Correlated spontaneous activity sets up multi-sensory integration in the developing higher-order cortex

Version 1

Abstract

Abstract

To perceive and navigate complex sensory environments, animals combine sensory information from multiple modalities in specialized brain circuits. Known as multisensory integration, this process typically depends on the existence of co-aligned topographic connections from several sensory areas to downstream circuits exhibiting multimodal representations. How such topographically co-aligned connectivity necessary for multisensory integration gets set up in early stages of development is still unknown. Inspired by the role of spontaneous activity in refining topographic connectivity between early sensory circuits, here we investigated the potential of such spontaneous activity to also guide the co-alignment of multiple sensory modalities in RL, a higher-order associative cortical area rostro-lateral to V1. Analyzing spontaneous activity simultaneously recorded in primary visual and somatosensory cortex and area RL at different developmental ages before sensory experience, we identify candidate features of this activity to guide the emergence of co-aligned topographic multisensory projections with somatosensory leading the visual projection. We confirm this hypothesis using a computational model of activity-dependent circuit refinement, and show that the correlation of spontaneous activity between the visual and somatosensory primary cortex can establish an optimal fraction of multisensory neurons in RL for stimulus decoding. Our model provides an exciting new computational perspective of the role of spontaneous activity in the emergence of topographically co-aligned multimodal sensory representations in downstream circuits, specialized for the processing of rich sensory environments.

Introduction

Sensory environments are complex – rich with objects and events that specific brain circuits must decipher to ensure their accurate perception and to guide appropriate behavior. To efficiently perceive and navigate sensory environments, these brain circuits combine information from multiple sensory modalities, including vision, somatosensation, or hearing, a process known as multisensory integration. Several brain regions have been shown to generate multimodal representations, typically by aligning connections from the sensory periphery where specific sensory features are first encoded to downstream projection areas. In the rodent, whiskers transmit tactile sensations which are combined with visual information to guide many behaviors including goal-directed behavior [1], gap crossing and food source detection [2], navigation [3], prey-capture behavior [4], and recognizing object features and shapes [5]. Multisensory integration has been reported in subcortical areas such as the superior colliculus [6, 7] and the striatum [8]. Higher-order areas (HOAs) situated between primary sensory areas have also been found to play a critical role in multisensory integration [1, 3, 9]. HOAs receive excitatory feedforward input via long-range axonal projections [10] and inherit and match cortical maps from primary sensory areas [1, 3, 11]. Moreover, connectivity is organized in a topographic manner across the modalities, whereby sensory neurons converge onto higher-order cortex while preserving space [3, 12]. Specifically, visual and tactile stimuli have overlapping representations in a higher order association area, area RL, that lies between the primary visual (V1) and the primary somatosensory (S1) cortices. There is significant spatial alignment between V1 and S1 in area RL; e.g. the lower whisker pad aligns to the lower visual field of view and can elicit responses in the same neurons, called bimodal neurons, in RL [3]. Additionally, this multisensory integration allows for cross-modal generalization. For example, rodents can adapt to changing sensory environments and can generalize sensorimotor tasks to a different modality, i.e. visual, after learning a task from only tactile stimulus, due to aligned abstract representations of these modalities in RL [1]. Finally, evidence shows that rats combine and integrate visual and tactile signals in a supralinear manner, significantly enhancing sensory processing efficiency [5]. Although full functional maturation of visual HOAs occurs only after eye opening, basic properties such as responsiveness to stimuli and topographic organization between V1 (as well as S1) and RL are already present at eye opening [13, 14].

Before the onset of sensory experience, primary sensory cortices generate spontaneous activity, which plays a pivotal role in their proper maturation in terms of cellular properties and activity-dependent connectivity refinements [15–18]. Across the sensory areas, a general developmental profile of spontaneous activity can be observed: early global and dense spontaneous events transition to more localized and sparse patterns, typical of adult networks [15, 17, 19–21]. By promoting the refinement of appropriate connectivity [22–24], these spontaneous activity patterns prepare the developing cortex for future sensory input [25–27], but at the same time are affected by changing circuitry [28]. Spontaneous activity has been characterized and shown to guide the establishment of hierarchical architectures in HOAs in the visual system before eye opening [29], but much less is known about its role in the development and topographic co-alignment of multiple sensory modalities in multisensory HOAs.

Whereas activity-independent mechanisms are known to set up the first cues instructing axons where to go [30, 31], activity-dependent synaptic plasticity mechanisms instructed by either spontaneous activity or sensory experience, are crucial in refining this initial imprecise connectivity [22, 32, 33]. Indeed, various forms of Hebbian synaptic plasticity, which promotes the potentiation of synaptic strength as a result of correlated pre- and postsynaptic activity, have been proposed to guide the activity-dependent refinement of circuit connectivity in development [28, 34–38]. Some experimental work has also measured Hebbianlike learning rules specific to development [39]. Activity-dependent mechanisms likely also contribute to the establishment and alignment of maps from different sensory modalities to downstream circuits, which has been mainly investigated in the rodent superior colliculus and avian optic tectum [32]. For example, how visual experience aligns the auditory and visual maps during development has been mainly studied by manipulations that cause the two maps to misalign [40–44]. This suggests that these same principles might apply not only to realign and maintain maps after manipulation, but also to the activity-dependent emergence of aligned representations in developing circuits.

Here, we combine computational modeling and calcium imaging experiments to investigate the role of spontaneous activity in the emergence of aligned topographic maps from two primary sensory cortices (visual, V1 and somatosensory, S1) to RL, an associative cortical area rostro-lateral to V1, and the emergence of bimodal responses in RL. We find that spontaneous activity in V1, S1, and RL sparsifies during development, but S1 exhibits more mature activity patterns compared to V1, which desynchronize as early as postnatal day eight. Using these experimental insights, we expand a computational model of activity-dependent connectivity refinement between primary sensory areas (V1 and S1) and RL using biologically plausible Hebbian synaptic plasticity which was previously used to study topographic map-formation and receptive field refinement between the thalamus and V1 [28]. Our model predicts that aligned topographic connectivity refinement and the emergence of bimodal neurons requires an intermediate amount of correlation between the primary sensory cortices (V1 and S1), which we investigate in our data. This connectivity refinement through spontaneous activity between the primary sensory areas and RL is sufficient to explain the developmental sparsification in RL whereby RL neurons become responsive to either or both primary sensory cortices in a ratio that ensures optimal information processing.

Results

Spontaneous activity in V1, S1 and RL during development

To investigate how spontaneous activity matures differentially across the mouse’s developing primary visual (V1), somatosensory (S1), and higher-order areas, specifically area RL, we performed two-photon imaging with one field of view per developing cortex in unanesthetized mouse pups at postnatal days (PN) 8-16 (Figure 1A,B). Spontaneous activity was observed across all three cortical areas (Figure 1C). We used a linear mixed model analysis [45] to estimate the differential effect of age on different aspects of spontaneous activity. As described in previous experiments [17, 21], we found that spontaneous activity in V1, sparsifies throughout development, i.e. the average amplitude, duration, and participation rate of spontaneous activity decrease with age, as reflected in the significant negative effect of age on these properties (Figure 1D-F, H, see Methods). In contrast, the frequency of events increases with age, as reflected in the significant positive relationship between age and event rate (Figure 1G, H). Similar trends of activity sparsification were also observed in S1 and RL. There were no differences between participation rate across areas, with all areas following a similar decreasing trend over time (Figure 1F,H). Furthermore, comparing area RL to V1 also showed no significant differences in duration of events and event rate between these two areas (Figure 1E,G,H), although amplitudes were slightly higher in RL compared with V1. However, we observed some significant differences, especially in S1. Specifically, spontaneous events in S1 have significantly lower amplitudes, lower durations, and higher rates compared to the other regions, at younger ages (Figure 1D,E,G,H). Together, this suggests that spontaneous activity in S1 has more mature amplitudes, duration, and event rates earlier in development, whereas RL and V1 develop more similarly during the second postnatal week. RL was found to slightly lag behind V1 and S1.

Spontaneous activity in RL sparsifies over development, with more mature activity in S1 compared to V1 and RL. A. Schematic of the two-photon experiments. Primary visual cortex (V1), somatosensory cortex (S1), and rostro-lateral area (RL) imaged in sequence over a 100 min recording session. B. Sample field of views of the individual cortices. Scale bar corresponds to 50 µm. C. Sample z-profile traces from the three different cortices (columns) in three mice at different postnatal ages (rows). Vertical scale bar corresponds to 20% ΔF/F0. Horizontal scale bar corresponds to 50 seconds. D-G. Amplitude (D), duration (E), participation rate (F), and event rate (G) as a function of postnatal age. Individual dots represent averages across all events in an area from one animal. Regression lines derived from a linear mixed model in H. H. Effect sizes of age and cortical region, and interaction effects of age within cortical region (i.e. Age:S1 and Age:RL) as determined by linear mixed model analysis, where age and region are included as fixed effects, and mouse ID and sample ID are included as random effects. All comparisons are relative to V1 at PN8. Values shown for V1 are estimates of the intercept at day PN8. Values shown for Age in V1 are estimates of the slope in the linear mixed model. All interaction effects of age within region (Age:S1 and Age:RL) apply to differences in the slope for that region compared to the slope of age within V1, whereas cortical region (S1 and RL) effects apply to differences in the intercept compared to V1 at PN8 (see Methods). Stars indicate statistical significance (* p<0.05, ** p<0.01, *** p<0.001). N=10 animals.

In summary, spontaneous activity in developing primary cortices and RL sparsifies with different temporal profiles, where the somatosensory cortex leads before the visual cortex and RL.

The primary somatosensory and visual cortex exhibit correlated spontaneous activity

Because RL needs to integrate activity from both V1 and S1 simultaneously, we hypothesized that there could be correlated activity between the primary sensory cortices to guide the emergence of this multisensory integration via some activity-dependent mechanism. To this end, we reanalyzed our published wide-field calcium imaging recordings from both V1 and S1 [46] to assess the temporal relationship between spontaneously occurring activity in these two primary sensory cortices (Figure 2A). We found that activity between the visual and the somatosensory cortex is often, but not always, temporally synchronized (Figure 2B). Indeed, analyzing the average activity between V1 and S1 of several animals at postnatal days PN9-PN12 revealed a range of Pearson correlation coefficients, with a mean of ∼0.5 (Figure 2C).

Spontaneous activity in S1 and V1 is synchronized. A. Schematic of wide-field experiments. An area encompassing the visual (V1) and the somatosensory (S1) cortices imaged over a 100 min recording session, broken up into 5 min recordings. B. Top: Three examples of spontaneous events, predominantly in V1 (left), S1 (right), or in both primary cortices (middle). Bottom: Sample ΔF/F0 traces averaged over V1 (red) or S1 (blue). Lines indicate the time points at which the events in the top row occurred. Horizontal scale bar indicates 20 seconds. Vertical scale bar represents 50% ΔF/F0. C. Temporal correlations between V1 and S1. Violin plots where each dot indicates Pearson correlation coefficients during 2-minute bin. PN9 N=6; PN10 N=5; PN11 N=4; PN12 N=2. D,E. Spatial correlations in V1/RL and S1/RL. Functional correlation maps show that spontaneous activity in RL is coordinated topographically with activity in both V1 (D) and S1 (E). Each color represents the correlations between the fluorescent changes over time of one of three seed areas in each image (white squares) and all pixels in the field of view (for details see Methods). Higher-order areas of the visual cortex are drawn according to [11]. LM: lateromedial area, AL: anterolateral area, A: anterior area.

Previous work has shown that connectivity between V1 and RL as well as S1 and RL is topographically organized in mature animals to preserve space [3, 11]. Furthermore, this organization is also aligned across the modalities, i.e. stimuli in the same region of sensory space tend to converge onto the same cells in RL [3, 12, 29]. Hence, in addition to temporal correlations, we evaluated whether spontaneous activity also carries information about spatial topography between the two primary sensory cortices and RL. We computed functional correlation maps for three distinct locations in V1 and RL, as well as three other locations in S1 and RL, showing that spontaneous activity in RL is coordinated topographically relative to both V1 and S1 (Figure 2D,E). Thus, spontaneous activity between V1 and RL and between S1 and RL is also spatially correlated and carries topographic information. This suggests that some activity-dependent mechanism might be using these temporal and spatial correlations during development to establish topographic connectivity maps between the primary sensory areas and RL, and hence enable multisensory integration in the adult animal.

Correlated spontaneous activity can generate topographic connectivity between the primary sensory cortex and RL and bimodal cells in RL

To test the hypothesis that an activity-dependent mechanism may use the spatio-temporally structured spontaneous activity in the primary sensory cortices to set up topographic connectivity from each area into RL and generate bimodal cells, we built and analyzed a computational model with three populations of neurons for V1, S1, and RL (Figure 3A) [28]. We modeled the activity-dependent synaptic plasticity of feedforward connections between V1 and RL, and S1 and RL, with each area subject to spontaneous activity as characterized before (Figure 1). While we initialized the connectivity as random, we included a slight bias for neighboring cells in the primary sensory cortices to project onto neighboring cells in RL (Figure 3B). This bias incorporates activity-independent effects like molecular gradients that are known to set up a coarse topography before the onset of activity-dependent plasticity, at least for connectivity between the sensory periphery and subcortical or primary cortical areas [38, 47–49]. Activity-dependent plasticity was implemented through a synaptic plasticity rule involving Hebbian and heterosynaptic terms. The Hebbian term potentiates synaptic weights based on coincident pre- and postsynaptic activity. The additional heterosynaptic term was implemented to depress synaptic weights, but only in the presence of presynaptic activity (see Methods)

Model set-up and assumptions. A. Schematic of a three-population feedforward model of the developing V1, S1, and RL. B. Example of random initial connectivity matrix between V1 and RL (connectivity matrix for S1 and RL is chosen identically) with a topographic bias denoted by stronger connections along the diagonal inspired by activity-independent mechanisms (see Methods). C. Sample of spontaneous events in V1 (top) and S1 (bottom) across cells as a function of time. Boxes highlight two spatiotemporally correlated events in V1 and S1 (see Methods). D. Example of refinement of connectivity matrix between V1 and RL (top) and S1 and RL (bottom), from t0 to tN. Same axes as in Figure 2B. E. Example of connectivity matrices in steady state from simulations with low (left) to highly (right) temporally correlated activity between V1 and S1. F. Schematic to illustrate topography and alignment measures (see Methods). G-I. Topography (G), alignment (H), and percentage of bimodal cells (I) as a function of V1 to S1 correlation. Red dots correspond to the examples from E. K. Excerpt of activity in RL at the beginning (left) and end (right) of the simulation (corresponds to panel D, left and right).

This rule was previously derived from interacting molecular neurotrophins [39, 50, 51] and shown to produce selectivity to correlated inputs from random initial connectivity [37, 52]. We then implemented spatiotemporally correlated spontaneous activity in areas V1 and S1 as measured experimentally (Figure 2C) which this synaptic plasticity rule used to modify the synaptic weights in the network. Specifically, to implement temporal correlations, S1 and V1 neurons were driven by a combination of events, independent and shared in each V1 and S1. To implement spatial correlations within a given area (V1 or S1), each independent event activated a randomly chosen, contiguous set of neurons, which we have previously shown can establish topographic connectivity between two areas [28]. To implement spatial correlations between V1 and S1, the same set (in terms of topographic identity) of neurons in both S1 and V1 were activated. The common inputs giving rise to shared events might reflect shared inputs from the subcortical areas such as the thalamus or superior colliculus into the primary sensory cortices [53–55] or correlation induced by direct connections between V1 and S1 [56, 57].

Using this spatiotemporally correlated spontaneous activity in the primary sensory cortices, the Hebbian plasticity rule (Eq. 1) refined and stabilized connectivity between V1/S1 and RL (Figure 3D), as shown before in a two-layer network [28]. We investigated the emergent connectivity properties for a range of V1 and S1 temporal correlations (Figure 3E). To quantify the topographic organization of connectivity between V1 and RL, and S1 and RL, and the alignment of maps from V1/S1 to RL, we defined two mathematical measures, topography and alignment (Figure 3F, see Methods). Weak temporal correlations (where independent events dominate over shared events) resulted in strong competition between the primary sensory cortices with dominant synaptic weight depression. The strong competition generated sparse connectivity matrices to RL, with many RL neurons decoupling from the primary sensory cortices (Figure 3E, left). This resulted in poor topography and map alignment of the maps between V1/S1 and RL (Figure 3G,H). RL had almost no bimodal cells, i.e. neurons that receive input from both primary sensory cortices, in contrast to what is found in mature RL experimentally [3] (Figure 3I). Increasing temporal correlations between the primary sensory cortices increased alignment of the V1/S1 to RL maps (Figure 3H). However, very strong correlations were detrimental to V1/RL and S1/RL topography (Figure 3E,G). Strong temporal correlation between V1 and S1 rendered many RL neurons selective to visual and somatosensory inputs at the same topographic location as reflected in the identical connectivity matrices (Figure 3E, right), overriding the initial topographic bias (Figure 3B). Only a medium-strength temporal correlation between V1 and S1, as found experimentally (Figure 2C), generated a connectivity profile which prevented RL cells to decouple from V1 and S1, while preserving topography (Figure 3E, middle).

We quantified the percentage of bimodal neurons in our model as a function of the temporal correlation between V1 and S1 (Figure 3I). By performing a steady-state analysis of our model (see Methods), we derived the critical amount of temporal correlation, c∗, between S1 and V1 required for the emergence of bimodal neurons. This parameter depends solely on two parameters of the model, i.e. the average firing rate of cells in the sensory cortices, ⟨pre⟩, and the heterosynaptic offset from Eq. 1, With these connectivity refinements, the generated activity in RL became sparser in terms of amplitude and participation rate (Figure 3J), consistent with the sparsification observed in the experimental data (Figure 1C,D,F).

In summary, our proposed computational model can account for the developmental sparsification of spontaneous activity in higher-order cortical area RL through the refinement of feedforward connectivity between two primary sensory cortices, V1 and S1, and RL. Furthermore, the model predicts that only medium-strength correlations between primary cortices produce topographically organized and aligned connectivity in V1/RL and S1/RL, and a mixture of bimodal and unimodal neurons in RL consistent with experimental data.

More mature spontaneous events in S1 can guide map alignment between V1 and RL

Having established that the spatio-temporal correlations between V1 and S1 can set up topographic and aligned connectivity to RL as well as generate bimodal cells in RL, we next turned to investigate a possible role of the differentially more mature spontaneous activity in S1 compared to V1 (Figure 1H). To examine this, we systematically increased the bias of the topographic connectivity between S1 and RL (Figure 4A). While unsurprisingly a stronger bias improved the topographic connectivity between S1 and RL (Figure 4B, right), it also improved the topographic organization of connectivity between V1 and RL (Figure 4B, left, and C). Therefore, spontaneous activity in S1, which matures more rapidly than V1, can guide the refinement of connectivity between V1 and RL.

More mature spontaneous events in S1 effectively instruct map alignment between V1 and RL. A. Three pairs of connectivity matrices (V1 left, S1 right) with increasing (top to bottom) S1 bias. B. Corresponding connectivity matrices in steady-state after plasticity. C. S1 and V1 topography for different S1-RL connectivity biases (indicated by color).

A mixture of bimodal and unimodal neurons in RL performs optimal decoding of activity

RL is involved in integrating information from different sensory modalities to obtain a more accurate representation of statistically rich sensory environments. In particular, an area that integrates visual and tactile stimuli is able to more accurately infer such information; e.g. when rats use both tactile and visual information, their performance in a task involving determining object orientation is better than the sum of the performance of each modality individually [5]. To investigate how the topographically aligned cortical connectivity between V1 and RL, and S1 and RL, might facilitate the integration of information from different sensory modalities, we quantified how much information about the activity of each primary sensory cortex can be decoded from the activity in RL. We quantified the amount of decoded information as the coefficient of determination, R2, which can be interpreted as the fraction of variance in V1 or S1 that can be reconstructed from activity in RL. To do this, we fitted a linear regression model using RL activity and both V1 and S1 activity, and calculated R2 from the error between our fitted linear model and each sensory area individually, which we refer to as the ‘V1 or S1 reconstruction’ (Figure 5A). [58, 59] (See Methods). R2 is naturally affected by how much activity in V1 or S1 is propagated into RL. In our model (Figure 3) and in biology [11, 60], some regions of the primary sensory cortices are not connected to RL; for example, RL contains only a partial representation of the visual field, hence V1 cells outside of that region do not project to RL [10,61]. Consistently, we found that the fraction of variance reconstructed, R2, for either V1 or S1 independently increased with the percentage of connected neurons from the respective cortex into RL (Figure 5B, see Methods). We showed analytically that a locally connected network, where every primary sensory cortical neuron makes at least one connection to RL and every RL neuron receives at least one connection from the primary sensory cortices, can maximize the total fraction of variance reconstructed for both V1 and S1 if the activity in V1 or S1 is perfectly temporally correlated (correlation = 1) (see Methods). These conditions translate to the scenario where all neurons in RL are bimodal. However, when activity across V1 and S1 is moderately correlated, having some unimodal RL neurons can achieve a higher total maximum fraction of variance for both V1 and S1 compared to the purely bimodal case (Figure 5C, see Methods). Having unimodal neurons increases variance explained when activity is not perfectly correlated because incomplete information from one modality can be completed with information from the other modality. Furthermore, the activity of unimodal neurons provides unambiguous information about whether only one or both of the primary sensory cortices is active, whereas in the case with all bimodal neurons this is indistinguishable. This suggests that the mixture of unimodal and bimodal neurons in RL [3] is the optimal solution for encoding the maximum amount of information from the activity in V1 and S1 [62].

A mixture of bimodal and unimodal neurons can optimally decode activity. A. Schematic of regression procedure to calculate R2 for reconstructing V1 or S1 activity from RL activity with a non-linear transfer function (a sigmoid) (see Methods). A linear regression is performed using the non-linearly transformed RL activity (dependent variable) and both V1 and S1 activity (independent variables). R2 is calculated from V1 or S1 activity separately using the fitted regression line. B. Reconstruction of V1 or S1 activity (R2) as a function of the percentage of connected cells from the respective cortex over all values of correlation chosen uniformly between 0 and 1 (see Methods). Color indicates topography. C. Reconstruction of V1 and S1 activity (R2) as a function of the percentage of bimodal cells (denoted by color).

Discussion

We investigated the role of spontaneous activity in the maturation of connectivity from two primary sensory cortices, V1 and S1, to RL, a higher-order association area, by combining computational modeling with experiments. We found that while spontaneous activity sparsifies during development in all investigated areas, S1 exhibits relatively mature, desynchronized activity patterns as early as postnatal day eight. Using these experimental insights, we constructed a computational model of activity-dependent connectivity refinement from each V1 and S1 to RL. To generate topographic connectivity from V1 and S1 to RL, aligned connectivity maps for the two sensory projections and a mixture of unimodal and bimodal neurons in RL, our model predicts that spontaneous activity between V1 and S1 should exhibit moderate temporal correlations, which we also found experimentally. Furthermore, we demonstrated how a more mature somatosensory cortex can guide the refinement of visual cortex connectivity to RL, suggesting that inter-cortical influence plays a central role in the establishment of refined and topographically aligned connectivity during development. Finally, we found that moderate correlations between V1 and S1 result in a mixture of bimodal and unimodal neurons in RL which can optimize information processing in RL when either one modality or both are active. Thus, by combining modeling with experiments, we found that activity-dependent plasticity can instruct the formation of refined and topographically aligned connectivity between primary sensory and high-order cortices in development to produce efficient information integration from multiple modalities.

Origin of correlated events in the primary sensory cortices

We find that early spontaneous activity with medium correlations can lead to the optimal amount of bimodal neurons and alignment of these sensory maps before sensory experience, which is crucial for effective multisensory integration in higherorder cortical areas. Such medium correlations indeed exist in the experimental data (Figure 2C). This result makes the prediction that alterations in the correlations in early spontaneous activity might affect later multisensory integration, which can be empirically tested. For instance, spontaneous activity in V1 has been shown to have different correlation properties in a mouse model of Fragile X [63] and in cases of disrupted inhibition [19, 46]; our model therefore suggests that multisensory integration in RL may be perturbed in these cases.

What might be the origin of the correlated activity we observed in the visual and somatosensory cortex (Figure 2)? One possibility is that the source is external to the animal [64], i.e. a single stimulus with both visual and somatosensory aspects. Indeed, while a substantial portion of activity in the developing cortex occurs independently of sensory stimulation [23], both the immature visual [65] and somatosensory [66] cortices exhibit some degree of sensory-driven activity. Although it is unlikely that this external stimuli gives any topographic information before eye opening, it could help to correlate the activity between sensory cortices. Alternatively, subcortical areas like the (primary or higher-order) thalamus are in an ideal position for coordinating activation across distinct regions of the cortex [29, 54, 67]. The developing thalamus exhibits spontaneous propagating waves, which have been shown to guide the formation of sensory maps to the cortex and support circuit plasticity [54,68,69]. Spontaneous calcium waves propagate among the different sensory-modality thalamic nuclei up to the cortex and hence provide a means of communication among sensory systems [69]. Finally, correlated events in V1 and S1 could also originate from lateral long-range connections between the visual and somatosensory cortex. These types of connections emerge during the first two postnatal weeks in mice [70] and potentially directly propagate activation from one area to another [56, 57]. More specifically, it has been shown that postsynaptic excitatory neurons in a subregion near the anterior border of V1, situated in close proximity to S1, receive input from local fast spiking inhibitory neurons that have direct cortico-cortical input from the posterior area of S1 [71]. This cross modal suppression may contribute to these medium level correlations during development.

Precise timing of map formation

We have shown that the somatosensory cortex’s early maturation is able to drive the aligned topographic development of the visual cortex Figure 4. This agrees with previous work showing earlier development of somatosensory cortex. For example, layer 4 (L4) to L2/3 excitatory synapses in somatosensory cortex were shown to mature 2 weeks earlier than those in visual cortex [72]. A similar template for the maturation of one map provided by another is observed in the refinement of cortico-collicular connections. Projections from S1 to the superior colliculus arrive as early as postnatal day 2 while projections from V1 enter the colliculus around postnatal day 6 [73], giving more time for S1 maps to refine. The precise timing of connectivity refinement between S1 and the superior colliculus likely guides the later developmental connectivity refinement between V1 and the colliculus and helps align the two maps [73, 74]. Moreover, this process of visual and somatosensory map alignment in the superior colliculus seems to be activity-dependent, with spontaneous activity in the retina playing an important role by driving the alignment of cortico-collicular connections from V1 to a genetically-induced duplicated retino-collicular map [74]. This modality difference in map refinement with the somatosensory leading the visual map seems to be present also in the refinement of connections between S1 vs. V1 and RL as we have shown earlier. Although the activity-dependent plasticity mechanisms behind this are experimentally unknown, inspired by earlier works [28, 50], our models suggest that similar activity-dependent plasticity involving Hebbian and heterosynaptic terms are sufficient.

Role of recurrent connectivity

Neural dynamics in the adult cortex are strongly affected by recurrent connections [75, 76]. In our model, we have chosen to focus on feedforward connections between the primary sensory and RL for several reasons. First, recent experimental evidence from the developing somatosensory cortex suggests that feedforward connectivity might be established first [77], and second, the strength and functional specificity of recurrent connections in the visual cortex increases only later in development [78]. Furthermore, abstraction is important for deriving robust biological insights from models [79] and here we are focused on how these feedforward connections arise and refine themselves over development building on our previous work of refinements between the thalamus and V1 [28].

Uni- and bimodal neurons in other species

Neurons responding to only a single or to multiple modalities are common in other animals apart from mice, like macaques [80–82], cats [83–85], or ferrets [86, 87]. The surprising presence of unimodal neurons in association cortices has inspired a number of studies attempting to explain their functional significance [85, 88, 89]. Close examination of the receptive fields of neurons in the associative cortex of the cat revealed that stimulus location is likely represented in a distributed population code and that unimodal neurons could contribute by improving the accuracy of the encoded stimulus location [85]. Furthermore, the response patterns of unimodal neurons in the cat extrastriate cortex revealed that while these neurons do not respond to their non-preferred modality, their activation is nonetheless modulated when non-preferred and preferred stimulus occur at the same time [88], further suggesting that unimodal neurons contribute to a population code. Varying the proportion of uni- and bimodal neurons might thus contribute to a continuum of multisensory responses across different cortical areas [90]. Thus, our proposed explanation for the mixture of uni- and bimodal neurons in terms of optimal decoding of input origin integrates well into the existing literature.

Methods

Experimental procedure Animals

All experimental procedures were approved by the institutional animal care and use committee of the Royal Netherlands Academy of Sciences. C57BL/6J mice of both sexes were used. All animals were aged between postnatal days (PN) 8-16. Mice of this strain open their eyes at P14.

In utero electroporation

For wide-field imaging (Figure 2) pyramidal neurons in layer 2/3 of the visual cortex were transfected with GCaMP6s (2 mg/ml) and DsRed (2 mg/ml) at embryonic day (E) 16.5 using in utero electroporation [91]. Pregnant mice were anesthetized with isoflurane and a small incision (1.5–2 cm) was made in the abdominal wall. The uterine horns were carefully removed from the abdomen, and DNA was injected into the lateral ventricle of embryos using a sharp glass electrode. Voltage pulses (five square wave pulses, 30 V, 50-ms duration, 950-ms interval, custom-built electroporator) were delivered across the brain with tweezer electrodes covered in conductive gel. Embryos were rinsed with warm saline solution and returned to the abdomen, after which time the muscle and skin were sutured.

Surgery

Animals were anesthetized with isoflurane (3 percent in 1 l/min O2). After anesthesia had become effective, lidocaine was used for local analgesia and a head bar with an opening (Ø 4 mm) above the visual cortex (0.5–2.5 mm rostral from lambda and 1–3 mm lateral from the midline) was attached to the skull with superglue and dental cement. For calcium imaging, a small craniotomy above the visual cortex was performed. The exposed cortical surface was kept moist with cortex buffer (125 mM NaCl, 5 mM KCl, 10 mM glucose, 10 mM HEPES, 2 mM MgSO4 and 2 mM CaCl2 [pH 7.4]).

Calcium indicator application

For the imaging experiments shown in Figure 1, the calcium-sensitive dye Cal-590 (Aat Bioquest) was dissolved in 4 µl pluronic F-127, 20 percent solution in DMSO (Invitrogen) and further diluted (1:10) in dye buffer (150 mM NaCl, 2.5 mM KCl, and 10 mM HEPES) to yield a final concentration of 0.5 mM. The dye was then pressure-ejected into S1, RL and V1 at 10–12 psi for 8–13 min with a micropipette (3–5 MΩ) attached to a picospritzer (Toohey).

Image acquisition

One hour before imaging, anesthesia was stopped, and animals were recorded in the absence of anesthesia. In vivo calcium imaging with cellular resolution (Figure 1) was performed on a Nikon multi-photon microscope (A1R-MP) with a 0.8/16x water-immersion objective and a Ti:Sapphire laser (Chameleon II, Coherent) Nikon software. We recorded the movement signal of the scan mirrors to synchronize calcium imaging and electrophysiology. Pixel size was 650 nm and images of 330 by 330 µm were recorded at 5-10 Hz.

In utero electroporated pups were used for wide-field calcium imaging of visual and somatosensory cortex (Figure 2). Calcium events were recorded with a Movable Objective Microscope (MOM, Sutter Instrument). Time-lapse recordings were acquired with a 4x objective (0.8 NA, Olympus) and blue light excitation from a Xenon Arc lamp (Lambda LS, Sutter Instrument Company). A CCD camera (Evolution QEi, QImaging) was controlled by custom-made LabVIEW (National Instruments) based software and images were acquired at a frame rate of 20 Hz.

Image processing

Images were processed as described previously [46]. For two-photon image processing, drift and movement artifacts were removed from each recording using NoRMCorre [92]. Each recording was aligned to the first recording in the series to remove any movements between recording sessions. Delta F stacks were made using the mean fluorescence per pixel as baseline, F0. ROIs were hand-drawn using ImageJ (NIH). Automated transient detection and further data processing was performed using custom-made MATLAB software (MathWorks).

Epifluorescence image processing: Delta F stacks were made using the average fluorescence per pixel as baseline. V1 was identified based on activity coordinates and shape after this method of identification was confirmed through immunohistochemistry for vGluT2 [19].

Statistical analysis

In two-photon recordings (Figure 1), activations within each cell were detected as in [93] as a peak with both an absolute height and relative prominence of at least 5% ΔF/F0. When more than 20% of recorded cells were active simultaneously, the activation was considered an event. In wide-field calcium imaging (we re-analyzed data from [46] (Figure 1)), events were detected as calcium peaks with an amplitude above 2x the standard deviation.

Functional correlation maps

Functional correlation maps were generated as in [29] as follows: Three seed areas were chosen (10x10 pixels each) within V1 or S1 and their time courses across a recording were determined by averaging the pixel values in each seed area for each frame of the recording. Next, the Pearson coefficient of correlation between the time courses of each seed area and those of all pixels in the field of view were determined and represented in one image for each seed area. Finally, the three images of each seed area were combined into one RGB image, such that each channel represents one functional correlation map. The actual color of each pixel represents the highest coefficient of correlation value across the three channels. The final RGB map reflects the topographic organization of the recorded spontaneous activity patterns.

Linear mixed model

To correctly account for the nested data structure commonly encountered in neuroscience research [45], we fit a linear mixed-effects model that treats animal and sample identity as random factors. We included interaction effects between age and region within our fixed effects. No transformations of the data were performed. Estimated effect sizes are shown in (Figure 1). Effect sizes on cortical region affect the intercept, but age and age-region interaction estimates affect the slope of the regression line. One-hot encoding was used because the regions are categorical variables. This means that one region, i.e. V1, is chosen as baseline and every other region is compared to this baseline. Any differences due to region are added/subtracted from the intercept for V1. Age and age-region interaction effects change the slope and should therefore be added or subtracted from the estimated slope for region V1. Since PN8 was the earliest age used in our data, 8 was subtracted from all ages so that PN8 is the estimated intercept. Our linear model may not be valid outside of the ages used (PN8-PN16) and no extrapolation should be done without experimental confirmation. This multilevel analysis allowed us to examine the effects of age and cortical region on four response variables: amplitude, event rate, participation rate, and duration. To obtain p-values, we used the Satterthwaite’s degrees of freedom method, as implemented in the lmerTest package [94] in R. The linear mixed model formula is: where y is one of the response variables, i.e. amplitude, event rate, participation rate, or duration and age is postnatal day. β0 is the intercept for V1 at age PN8, and β1 and β2 are the differences in intercept for cortical region (S1 and RL are dummy variables set to 0 if not predicting that region or 1 if predicting that region). β3 is the slope for V1 over age and β4 and β5 are the differences in slope (age-region interaction effects) from V1 over age for the cortical region of interest. ɛij is the error from both random effects (mouse ID and sample ID) and residual error.

Rate-based network model

Extending the model introduced by [28], we studied a feedforward, rate-based network with two onedimensional input layers, v and s, and one output layer r with N v, N s, and N r neurons each. We used periodic boundary conditions in each layer to avoid edge effects. The initial connectivity between v and r, Wv, and s and r, Ws, is initially all-to-all with uniformly distributed weights in the range wini ∈ [a, b]. In addition, a topographic bias was introduced by modifying the initially random connectivity matrix to have the strongest connections between neurons at the matched topographic location. These biased weights decay with a Gaussian profile with increasing distance (Figure 3)), with amplitude a and spread s. During the evolution of the weights, soft bounds were applied on the interval [0, Wmax].

The input populations v and s receive spontaneous activation from a homogeneous Poisson process with an inter-event interval of 150s, implemented using a Binomial random variable with sufficiently small time steps. A spontaneous event increased the firing rate of a fraction of neighboring neurons to 1 for Ldur seconds. The activity of the output layer r is given as a leaky integrator of the input populations, We studied the weight evolution under activity-dependent learning rules. In particular, following [28] we chose a rule that combines Hebbian and heterosynaptic effects that can produce selectivity, and where θ is a heterosynaptic offset constant that defines a distance-dependent heterosynaptic depression. Heterosynaptic plasticity occurs when a nearby unstimulated synapse stabilizes or depresses due to the plasticity in a stimulated one. This helps to stabilize the Hebbian plasticity [50]. Table 1 lists all parameters. For simulations with varying correlation values, the % of correlated events is chose from a uniform distribution between 0 and 1, U[0,1]. For simulations with varying S1 bias, a Gaussian spread of 4 and a random bias from a uniform distribution between 0.05 to 0.6 was chosen. Simulation time of 500,000 ms was used.

Expected weight dynamics and bimodality threshold

Since the dynamics of the output layer are much faster than the change in weight, r << w, we made a steady state assumption and computed This allowed us to write the expected change in weight between (e.g.) the v and r populations as a function of the input statistics of the v and s population, This formulation allowed us to investigate the critical required amount of correlation between cells in v and s that allows for the formation of stable bimodal cells in r. In particular, we investigated the expected change of weight in the configuration where all connections between v and r are fully depressed and only one connection between s and r is potentiated: and The expected weight change in this configuration is where Cov is the covariance and µv, µs are the average firing rates. The critical amount of correlation, where potentiation and depression balance each other out, is characterized by the condition which we solved for the correlation, γsv = Cov(si, vi)/(σvσs), as Assuming that rates and variances in v and s are equal and that the activity is Poisson, µ = σ2, the condition for the critical correlation further simplifies to

Optimal weight matrix for correlated input populations

We considered a concatenation of the two “input” populations, X = [v; s] of size M × 1, that drive activation in the “output” population, r = ϕ(WX), where ϕ is a (possibly non-) linear transfer function, r is size N r × 1 and M = N v + N s. The decoding problem introduced in Figure 5 can be phrased as inferring activity of the input populations from the output population, = Aϕ(WX), where A is the pseudoinverse of W, A = (WT W)−1WT. We then searched for the optimal weight matrix, Wopt = [Wv, Ws] of size N r × M, where Wv represents the weights from v to r, and Ws represents the weights from s to r, that minimizes the squared error between X and , E = (X − )2 and therefore maximizes R2.

As the dimension of r is substantially smaller than the dimension of X, dim(r) ≪ dim(X), the optimal weight matrix needs to preserve the largest possible amount of variability of the input while projecting it into a lower dimensional space. Using insights from principal component analysis theory [95], we can demonstrate that the optimal weight matrix is composed of eigenvectors of the covariance matrix of X with the largest eigenvalues, To better understand the structure of the eigenvectors, we can exploit the fact that in our set-up the covariance matrix, C is circular, i.e. every row of XT X can be obtained by a cyclic shift of the generating vector, c = [c0, c1, …, cM−1]: Eigenvectors and eigenvalues of circulant matrices have well-known closed form solutions [96]. The eigenvectors are the Fourier modes: where ω = e2πi/M, a primitive M -th root of unity. The corresponding eigenvalues can be computed as Since the covariance matrix is symmetric, i.e. cj = cM−j for all j and all values are real in the covariance matrix we can rewrite the eigenvalues in terms of their real parts: For a real-symmetric circulant matrix, the real and imaginary parts of the eigenvectors are themselves eigenvectors:

The first M/2 eigenvectors with nonzero eigenvalues from v and w will comprise Wopt which maximizes R2. Only the first M/2 are used since the dimension of r is M/2 in our case where N v = N s and can have a maximum of M/2 eigenvectors.

Our experimental set-up with correlated input populations translates into a generating vector c that has nonzero terms around the diagonal and offset from the diagonal by M/2. If the inputs between V1 and S1 are fully correlated, then ck = 1 for k ∈ {0, . . . , W/2, (M − 1)/2 − W/2, . . . , (M − 1)/2 + W/2, (M − 1) − W/2, . . . , M − 1} and ck = 0 else, where W is the extent of spatial correlation of the neurons. Taking this into consideration and removing any terms that simplify to zero due to the periodicity of the cosine function gives us the eigenvalues: Hence, the eigenvectors with nonzero eigenvalue of the covariance matrix are exactly the first half of the eigenvectors of the sin and cos functions with an even-numbered period. Due to this simplification of the odd j’s leading to λj = 0 and even numbered periods, this results in eigenvectors that are identical on the indices {0, . . . , M/2 − 1} and {M/2, . . . , M } which is the definition of a bimodal cell and therefore all cells must be bimodal to maximize the R2 when input correlation = 1 between V1 and S1.

Topography

Network topography measures the degree to which the initial topographical bias is preserved in the final receptive field. To calculate topography, we calculated the Pearson’s correlation coefficient, r, of the final receptive field with an ideal receptive field template (Figure 3). The ideal receptive field template that was used has weights of 1 with a Gaussian spread of 4 from center.

Alignment

Alignment is a measure of similarity between the final receptive fields of V1 and S1. To calculate alignment, we calculated the Pearson’s correlation coefficient, r, between the final receptive field of S1 and the final receptive field of V1.

Reconstruction

To calculate the reconstruction of V1 and S1 activity from RL in Figure 5, we fitted a linear model between all 3 layers’ activity patterns for 1000 sec using the final weight matrix of RL (independent variable) and V1 and S1 combined (dependent variable). Duplicates of RL activity pattern were removed from all layers activity and 80% of the data was used to avoid over-fitting. Then the coefficient of determination was calculated as where SSres is the sum of the residuals squared and SStot is the total sum of squares, using only one sensory area’s activity at a time, e.g. RL vs. V1 or RL vs. S1 activity compared with the linear fit of all 3 layers.

eLife Assessment

This valuable study combines experiments and theory to investigate the putative role of spontaneous correlated activity in establishing aligned topographic maps of neural activity in higher-order sensory areas, and will be of interest to researchers studying multisensory integration and brain development. However, the evidence presented is incomplete, as there are notable disconnects between the experimental data and the modeling setup, and there are methodological details that are either unclear or missing, limiting the strength of the claims.

Reviewer #1 (Public review):

Dwulet et al. combined experimental and modeling approaches to investigate how correlated spontaneous activity in the mouse's primary visual (V1) and primary somatosensory (S1) areas drives the development of multisensory integration in area RL. Notably, they focused on early developmental stages, before sensory experience occurs. Consistent with previous experimental findings, the authors first demonstrated that spontaneous activity becomes more sparse across development in all three areas, as measured by event amplitude, event duration, and participation ratio. Using a linear mixed model analysis to compare the maturation of this spontaneous activity, they found evidence that S1 matured the fastest. The authors then presented experimental evidence suggesting that these spontaneous events were moderately correlated both spatially and temporally.

They hypothesized that activity-dependent mechanisms use these correlations to establish connectivity across these regions. To test this hypothesis, the authors modeled a feedforward network with connections from S1 to RL and from V1 to RL, where the strength of connections depended on a Hebbian term for potentiation and a heterosynaptic term for depression. By investigating different levels of V1-S1 correlations, they found that moderate levels of correlation led to the significant development of topographically organized connectivity while maintaining a mix of bimodal and unimodal cells in RL. Additionally, when simulating a network with a more mature S1, they observed that topographical maps improved not only between S1 and RL but also between V1 and RL. Finally, the authors use linear regression to suggest that the mixture of bimodal and unimodal cells in RL is optimal for encoding the maximum amount of information from both V1 and S1.

However, there are significant gaps between the experimental data and the modeling setup, which weaken the paper's conclusions. Additionally, some key details are omitted, making it difficult to fully assess their analysis and interpret some of their figures.

(1) Some of the statistical measures and techniques in Figure 1 could benefit from clearer definitions. While the thresholds for activation (peak with at least 5% dF/F0) and events (20% of recorded cells activated simultaneously) are provided, event duration and participation rate are not clearly defined. Based on this definition of event alone, it is unclear why the minimum participation rate in Figure 1F is not 20%. Additionally, the conclusion that S1 matures earlier than RL and V1 could be strengthened by including a direct comparison between S1 and RL, as the current analysis only compares these areas to V1.

(2) The wide-field experiments in Figure 2 could be expanded to support the feedforward modeling assumptions. Currently, the spatial and temporal correlations presented leave open the possibility that these spontaneous events are traveling waves propagating from V1 to RL to S1 (or vice versa). This scenario would suggest a different connectivity scheme for the model. Clarifying this point with additional data analysis, specifically including temporal correlations involving RL, could provide stronger support for the model's assumptions.

(3) The functional correlation map in Figure 2D appears contradictory to the authors' modeling assumption that inputs are correlated spatially in V1 and S1. While V1 seed points align topographically with RL, this organization breaks down when extended into S1. In contrast, and in support of the modeling assumption, Figure 2E shows clearer topography across all three regions. A discussion of this discrepancy would be helpful, as it's a key conclusion of the figure. Additionally, it is unclear when this data was collected during development. Clarifying the developmental stage and analyzing how this map changes over time could strengthen the results.

(4) The modeling of spontaneous events with fixed amplitude and duration seems inconsistent with the experimental data in Figure 1, which shows variability in these parameters. This is particularly confusing in Figure 4, where S1 maturation is modeled as a stronger topographical alignment with RL, but the experimental data defines maturation based on amplitude, duration, and event rates. Justifying these modeling choices or adapting the model to reflect experimental variability would create a better connection between the theory and data.

(5) Several important details of the mathematical model are missing or unclear, partly due to typos. The Results section mentions the general framework of the input correlation matrix (e.g., "S1 and V1 neurons were driven by a combination of events, independent and shared in each V1 and S1" and "each independent event activated a randomly chosen, contiguous set of neurons"), but the specifics are not fully explained. Additionally, the caption of Figure 5 refers to a non-linear transfer function (a sigmoid), but these details are not provided in the Methods section, which instead suggests a linear model was used. A careful review of the main text and Methods section would help ensure that all the necessary details are included and that the story is both complete and accurate.

(6) While Figure 5 supports the paper's conclusion that a mixture of unimodal and bimodal neurons in RL optimizes information encoding, the authors missed an opportunity to strengthen the connection between the model and experimental data. Specifically, they could apply this reconstruction method to the experimental data and examine how RL's ability to reconstruct V1/S1 activity changes across development. Their model predicts that this performance would improve over time, and if this trend is observed in the experimental data, it would provide strong validation that these feedforward connections are developing in line with the model's predictions.

Reviewer #2 (Public review):

The authors aim to investigate the role of spontaneous activity in shaping the development of multisensory integration in the brain, specifically focusing on the connections between primary visual and somatosensory sensory areas (V1 and S1) and a higher-order cortical area rostro-lateral to V1 (RL). They seek to understand how spontaneous activity guides the formation of aligned topographic maps and the emergence of bimodal neurons in RL.

First, the authors found that spontaneous activity in all three areas sparsifies over time, but S1 exhibits more mature patterns earlier than V1 and RL. They claimed that correlated activity among neighboring regions of these areas during development carries topographic information. These data were used to implement a computational model that employed Hebbian rules of synaptic plasticity. The model indicated that correlated spontaneous activity can generate topographic connectivity between S1/V1 and RL and bimodal neurons in RL. The model suggested that the more mature spontaneous activity in S1 can guide map alignment between V1 and RL. In addition, the model also suggested that a mixture of bimodal and unimodal neurons in RL is optimal for decoding information from V1 and S1.

While the data presented in the manuscript is promising and provides preliminary insights into the role of spontaneous activity in multisensory integration, it would be beneficial to strengthen the experimental foundation regarding the correlation between V1, S1, and RL. Incorporating more rigorous spatio-temporal analyses of spontaneous activity could enhance the robustness of these findings.

Here are some important concerns:

(1) The analysis of how spatial topography influences activity correlations in Figure 2 has several issues.

1a. While squares in V1 and S1 covered a small area of these sensory areas, the correlated territories in RL covered the entire area of RL. The topographic map in V1 continues caudally, so where is the rest of the map in RL? Something similar applies to the relationship between S1 and RL.

1b. It is essential to know how areas were drawn. High precision is required.

1c. It is not clear if correlated activity means different events in sync or large events that cover 2 or all 3 cortical areas of interest. The figure points to the second option, which contradicts the size of events at these stages, mainly in the oldest mice analyzed here.

1d. It is fundamental to know in detail and provide examples of how the detection of events was performed. For instance, could the dispersion of light from an event in V1 close to RL cause the detection of activity in RL?

(2) For the correlations among V1, S1, and RL, it is crucial to have a consistent method to delineate the borders of cortical areas. The authors mention in one sentence that areas were drawn according to a reference map. More details are needed to convince the reader that the borders are accurate, especially because their shape and position change with age.

(3) The results from the model seem to be based on the initial bias in connectivity between neighboring cells from the different areas. Then, it seems straightforward that implementing correlated activity with Hebbian and synaptic depression rules will force the strengthening of connections between spatially close cells. Despite this apparent predisposition of the model towards a defined outcome, the flaws in the experimental data used prevent a rigorous interpretation of the computational model.

(4) In the Introduction, the authors nicely and briefly explain the role of primary and higher-order sensory cortices in information processing. They also explain how spontaneous activity during development helps to build these circuits by refining connections or establishing hierarchies. They continue explaining the relevance of aligning different topographic maps to allow multisensory integration. Then they provide some examples of sites of multisensory integration. This provides a general context for the data presented in the Results section; however, and importantly, there is no specific introduction of why they are interested in RL and its interaction with V1 and S1. The authors should introduce the RL area and explain why it is an interesting site for multisensory processing.

(5) The results shown in Figure 1 corroborate published data from Golshani et al, Rochefort et al, Murakami et al. While the reproduction of data is more than welcome, the authors should specify which part of the data is completely new and acknowledge clearly the rest as corroboration of previous data. The sentence "As described in previous experiments ..." partially acknowledges this fact but is not clear enough. In addition, the transition between this part of the manuscript and the next data is not smooth. Data seems to be used to feed the model so perhaps the organization of the manuscript leaves room for improvement.

Reviewer #3 (Public review):

Summary:

The study by Dwulet et al. explores how the development of spontaneous neural activity in primary sensory cortices influences the co-alignment of multiple sensory modalities in higher-order brain areas (HOAs). To address this question, they focus on connectivity between the primary visual (V1) and somatosensory (S1) cortices and an associative cortical area (RL) in mice. The authors combine experimental (wide-field and two-photon calcium imaging) and computational approaches to show that spontaneous activity matures at a different pace across these brain regions. Their data indicate that S1 develops more rapidly than V1, which is possibly beneficial for RL's integration of visual and somatosensory inputs through correlated spontaneous activity. Using a computational model, they demonstrate that a moderate correlation between V1 and S1 activity can optimally guide the formation of bimodal neurons in RL, which are crucial for maximizing the decodability of multisensory stimuli. This finding highlights the role of correlated spontaneous activity in primary sensory cortices in establishing co-aligned topographic multimodal sensory representations in downstream circuits.

Strengths:

The manuscript is well written and it provides strong enough evidence to support the main claim of the authors. The insights on the role of correlated activity on instructing co-aligned multisensory maps in HOAs are not trivial and are an important advancement for the field.

Weaknesses:

In the opinion of this reviewer, the study has no major weaknesses. A drawback of the work is that none of the predictions of the computational modeling have been corroborated through mechanistic experimental manipulations of early brain activity.

Author response:

Public Reviews:

Reviewer #1 (Public review):

Dwulet et al. combined experimental and modeling approaches to investigate how correlated spontaneous activity in the mouse's primary visual (V1) and primary somatosensory (S1) areas drives the development of multisensory integration in area RL. Notably, they focused on early developmental stages, before sensory experience occurs. Consistent with previous experimental findings, the authors first demonstrated that spontaneous activity becomes more sparse across development in all three areas, as measured by event amplitude, event duration, and participation ratio. Using a linear mixed model analysis to compare the maturation of this spontaneous activity, they found evidence that S1 matured the fastest. The authors then presented experimental evidence suggesting that these spontaneous events were moderately correlated both spatially and temporally.

They hypothesized that activity-dependent mechanisms use these correlations to establish connectivity across these regions. To test this hypothesis, the authors modeled a feedforward network with connections from S1 to RL and from V1 to RL, where the strength of connections depended on a Hebbian term for potentiation and a heterosynaptic term for depression. By investigating different levels of V1-S1 correlations, they found that moderate levels of correlation led to the significant development of topographically organized connectivity while maintaining a mix of bimodal and unimodal cells in RL. Additionally, when simulating a network with a more mature S1, they observed that topographical maps improved not only between S1 and RL but also between V1 and RL. Finally, the authors use linear regression to suggest that the mixture of bimodal and unimodal cells in RL is optimal for encoding the maximum amount of information from both V1 and S1.

However, there are significant gaps between the experimental data and the modeling setup, which weaken the paper's conclusions. Additionally, some key details are omitted, making it difficult to fully assess their analysis and interpret some of their figures.

(1) Some of the statistical measures and techniques in Figure 1 could benefit from clearer definitions. While the thresholds for activation (peak with at least 5% dF/F0) and events (20% of recorded cells activated simultaneously) are provided, event duration and participation rate are not clearly defined. Based on this definition of event alone, it is unclear why the minimum participation rate in Figure 1F is not 20%. Additionally, the conclusion that S1 matures earlier than RL and V1 could be strengthened by including a direct comparison between S1 and RL, as the current analysis only compares these areas to V1.

We thank the reviewer for this comment. We have now updated the Methods to include the event duration as time above half max, participation rate as % of cells out of total in that region active during an event. Also, the threshold of 20% recorded cells to identify an event was incorrectly stated, in fact the threshold was 5% consistent with what the reviewer observed in Figure 1F. This error has been corrected throughout the Methods. We chose 5% because spontaneous activity significantly sparsifies over development, with events involving far fewer cells, as previously shown by multiple studies (Golshani et al., 2009; Rochefort et al., 2009; Gribizis et al., 2019; Leighton et al., 2021; Murakami et al., 2022; Chini et al., 2022; reviewed in Lakhera et al., 2024).

For the linear mixed model (LMM) analysis, we used V1 as a reference just for convenience, but this has no influence on the results. We now added a direct comparison using each area as reference in the LMMs. Several Supplementary Tables (S1-3) now show these results with coefficient estimates and stars showing statistical significance and are mentioned in the legend of Figure 1 and the main text.

(2) The wide-field experiments in Figure 2 could be expanded to support the feedforward modeling assumptions. Currently, the spatial and temporal correlations presented leave open the possibility that these spontaneous events are traveling waves propagating from V1 to RL to S1 (or vice versa). This scenario would suggest a different connectivity scheme for the model. Clarifying this point with additional data analysis, specifically including temporal correlations involving RL, could provide stronger support for the model's assumptions.

We agree with the reviewer that the correlation analyses shown in Figure 2 do not differentiate between two possibilities: activity that travels smoothly from one cortical area to another, thereby correlating correlations between these areas, versus activity that is spatially confined to individual areas but occurs near-synchronously across those areas. To address this point, we have revised Figure 2 in two ways.

First, we added examples of spontaneous activity showing near-synchronous but spatially distinct activation of sub-areas in V1, RL and S1 (new Figure 2D). These examples show that localized activity can remain confined to individual sensory cortical areas and RL, while occurring at similar times across areas. Thus, the observed correlations are not simply due to single large events spreading continuously across the entire imaged field.

Second, we added a lagged cross-correlation analysis between V1 and S1 activity (new Figure 2G). This analysis shows that the correlation between V1 and S1 peaks close to zero lag and decays for both positive and negative lags. This argues against a stereotyped travelling-wave-like propagation from V1 to S1 or from S1 to V1 with a fixed delay. The cross-correlation curves show a mild asymmetry, with somewhat higher correlations when S1 precedes V1. However, because the dominant peak is centered near zero lag, we interpret the data primarily as evidence for near-synchronous, spatially structured coactivity across sensory areas, rather than fixed directional propagation.

Together, these two analyses support the modeling abstraction that V1 and S1 provide temporally correlated, spatially structured inputs to RL. We have added the new activity examples and the lagged cross-correlation analysis to Figure 2 and revised the Results accordingly. Although these analyses do not exclude all forms of propagating activity, they argue against the specific concern that the correlations are dominated by stereotyped traveling waves passing sequentially through V1, RL, and S1.

(3) The functional correlation map in Figure 2D appears contradictory to the authors' modeling assumption that inputs are correlated spatially in V1 and S1. While V1 seed points align topographically with RL, this organization breaks down when extended into S1. In contrast, and in support of the modeling assumption, Figure 2E shows clearer topography across all three regions. A discussion of this discrepancy would be helpful, as it's a key conclusion of the figure. Additionally, it is unclear when this data was collected during development. Clarifying the developmental stage and analyzing how this map changes over time could strengthen the results.

We thank the reviewer for pointing out this ambiguity. In the original version, the functional correlation maps were generated using separate seed locations in V1 and S1, and the interpretation relied heavily on thresholded RGB maps in which each pixel was assigned to the color channel with the strongest correlation. This representation made it difficult to directly compare the V1- and S1-seeded maps and may have given the impression that topographic organization was preserved in one direction but not the other.

We have therefore revised the analysis and presentation of Figure 2. Instead of using separate seeds in V1 and S1, we now use common seed locations in RL and compute the correlations of these RL seeds with activity across the imaged cortical field. This allows us to ask directly whether different RL locations are associated with spatially distinct regions in both V1 and S1. We now show both the raw correlation maps, in which the RGB channels reflect the correlation values for the three RL seeds (new Figure 2E), and the thresholded/maximum-channel representation, in which each pixel is assigned to the strongest of the three color channels (new Figure 2F). The raw correlation maps make the correlation structure visible without relying solely on thresholding, whereas the thresholded representation highlights the spatial ordering of the strongest correlations.

With this revised analysis, the topographic relationship across V1, RL, and S1 is clearer and no longer depends on comparing separate V1- and S1-seeded maps. We also clarified in the figure legend how the RGB maps are computed and how thresholded pixels are represented.

The reviewer also asked about the developmental stage and progression of this phenomenon. The example shown in Figure 2 was recorded at PN9, and we now state this explicitly. In addition, we added examples from PN9–PN13 in Supplementary Figure S1, showing that similar functional correlation-map structure is present across the developmental period analyzed here. This is consistent with previous work showing that retinotopy-like patterns in higher visual areas can be recovered from functional-connectivity analysis of spontaneous activity before eye opening (Murakami et al., 2022), and with recent work showing that retinotopy-like and somatotopy-like patterns of ongoing activity, together with their rough topographic correspondence in RL, are already present before eye opening at PN10–11 (Matsumoto, Murakami & Ohki, 2025).

(4) The modeling of spontaneous events with fixed amplitude and duration seems inconsistent with the experimental data in Figure 1, which shows variability in these parameters. This is particularly confusing in Figure 4, where S1 maturation is modeled as a stronger topographical alignment with RL, but the experimental data defines maturation based on amplitude, duration, and event rates. Justifying these modeling choices or adapting the model to reflect experimental variability would create a better connection between the theory and data.

We agree with the reviewer that the original presentation did not sufficiently distinguish between the experimentally measured maturation of spontaneous activity and the way S1 maturation was implemented in the model. In the experiments (Figure 1), earlier maturation of S1 was reflected by lower event amplitudes, shorter durations, and higher event rates. In contrast, the original model explored the effect of a stronger or more spatially refined S1-to-RL projection (Figure 4). This modeling choice was motivated by pilot anatomical data suggesting that projections from S1 to RL become more elaborate earlier than projections from V1 to RL at comparable developmental ages. We include examples of these pilot data (Author response image 1), but we have not included them in the manuscript because the dataset is preliminary and does not yet allow for a sufficiently complete quantitative analysis.

Projections from V1 and S1 to RL at different developmental ages.

Pilot anatomical data suggest that the S1 projection to RL becomes more elaborate and mature earlier than the V1 projection.

To address the reviewer’s concern more directly, we have now extended the model to incorporate differences in the spontaneous activity patterns of V1 and S1, including the lower amplitude and higher frequency of S1 events. We then examined how these activity differences interact with different levels of initial connectivity bias between the primary sensory cortices and RL (Supplementary Figure S2). We also quantified the resulting topography, map alignment, and fraction of bimodal RL neurons as a function of the S1 bias and included these additional plots in Figure 4 (panels C-E).

This analysis shows that incorporating the more mature S1-like activity patterns alone was not sufficient to generate the appropriate topographic and aligned maps. Rather, the model still required an initial connectivity bias, together with an appropriate level and structure of correlated activity. This is consistent with the results shown in Figure 3B,E,G–I and discussed in our response to Reviewer 2, point 3, where we show that the initial bias does not by itself determine the final map structure, but instead interacts with the level of V1–S1 correlation. We have added the new analysis to Supplementary Figure S2 and revised the text to clarify the interpretation. Rather than presenting the stronger S1 bias as a direct consequence of the more mature S1 activity dynamics revealed through the differences in amplitude, duration, and event rate, we now frame it as a model prediction: earlier S1 maturation may need to be accompanied by, or act through, a more advanced anatomical or functional S1-to-RL projection, whose refinement still depends on the temporal and spatial structure of spontaneous activity.

The results suggest that differences in spontaneous activity dynamics and differences in projection maturity may act together during the emergence of topographically aligned multisensory maps, with neither component alone being sufficient to determine the final organization. Future experiments will be needed to establish whether such an S1-to-RL connectivity bias is present systematically, to quantify its developmental progression, and to disentangle the relative contributions of more mature spontaneous activity dynamics and more mature connectivity.

(5) Several important details of the mathematical model are missing or unclear, partly due to typos. The Results section mentions the general framework of the input correlation matrix (e.g., "S1 and V1 neurons were driven by a combination of events, independent and shared in each V1 and S1" and "each independent event activated a randomly chosen, contiguous set of neurons"), but the specifics are not fully explained. Additionally, the caption of Figure 5 refers to a non-linear transfer function (a sigmoid), but these details are not provided in the Methods section, which instead suggests a linear model was used. A careful review of the main text and Methods section would help ensure that all the necessary details are included and that the story is both complete and accurate.

We thank the reviewer for pointing out these missing details and inconsistencies. We have carefully revised the Results, figure captions, and Methods to make the model description more complete and internally consistent.

First, we clarified how spontaneous input events were generated. Specifically, V1 and S1 activity was constructed from independent events in each area and shared events across the two areas. These event streams were generated using Poisson processes, with the rates chosen such that the total event rate was matched across simulations while varying the fraction of shared versus independent events. We also clarified that each event activated a spatially contiguous group of neurons, thereby implementing local spatial correlations within each primary sensory area, while shared events activated corresponding topographic locations in V1 and S1.

Second, in the Methods we clarified the use of the nonlinear transfer function in the decoding analysis shown in Figure 5. The simulated RL activity was transformed with a sigmoid nonlinearity before performing the regression analysis, and we have now added the corresponding equation (15) to the Methods.

Third, we clarified the distinction between the numerical decoding analysis and the analytical calculation of the optimal weight matrix. The decoding analysis uses the nonlinear transformation described above, whereas the analytical calculation uses a linearized version of the model to obtain a tractable closed-form solution. We now state this explicitly in the Methods to avoid the impression that two inconsistent models were used.

Finally, we corrected several typographical errors and checked that the Results, Methods, and figure captions use consistent terminology for the input generation, correlation structure, and decoding analysis.

(6) While Figure 5 supports the paper's conclusion that a mixture of unimodal and bimodal neurons in RL optimizes information encoding, the authors missed an opportunity to strengthen the connection between the model and experimental data. Specifically, they could apply this reconstruction method to the experimental data and examine how RL's ability to reconstruct V1/S1 activity changes across development. Their model predicts that this performance would improve over time, and if this trend is observed in the experimental data, it would provide strong validation that these feedforward connections are developing in line with the model's predictions.

We agree with the reviewer that applying the reconstruction analysis directly to the experimental data would provide an important additional test of the model. However, the current experimental datasets are not well suited for this analysis. The two-photon recordings used to characterize spontaneous activity in V1, S1, and RL were acquired sequentially rather than simultaneously, and therefore cannot be used to reconstruct V1/S1 activity from RL activity. In principle, a related analysis could be attempted using the wide-field recordings, which are simultaneous across cortical areas. However, these data have lower spatial resolution, include movement-related variability, and do not provide cellular-resolution measurements of RL activity. We explored this possibility, but the resulting reconstructions were not sufficiently reliable or interpretable to include in the manuscript.

We now state this explicitly as a limitation in the Discussion and identify simultaneous multiarea recordings at cellular resolution as an important future test of the model. Such experiments would make it possible to determine whether the ability of RL activity to reconstruct V1/S1 activity improves across development, as predicted by the model.

Reviewer #2 (Public review):

The authors aim to investigate the role of spontaneous activity in shaping the development of multisensory integration in the brain, specifically focusing on the connections between primary visual and somatosensory sensory areas (V1 and S1) and a higher-order cortical area rostrolateral to V1 (RL). They seek to understand how spontaneous activity guides the formation of aligned topographic maps and the emergence of bimodal neurons in RL.

First, the authors found that spontaneous activity in all three areas sparsifies over time, but S1 exhibits more mature patterns earlier than V1 and RL. They claimed that correlated activity among neighboring regions of these areas during development carries topographic information. These data were used to implement a computational model that employed Hebbian rules of synaptic plasticity. The model indicated that correlated spontaneous activity can generate topographic connectivity between S1/V1 and RL and bimodal neurons in RL. The model suggested that the more mature spontaneous activity in S1 can guide map alignment between V1 and RL. In addition, the model also suggested that a mixture of bimodal and unimodal neurons in RL is optimal for decoding information from V1 and S1.

While the data presented in the manuscript is promising and provides preliminary insights into the role of spontaneous activity in multisensory integration, it would be beneficial to strengthen the experimental foundation regarding the correlation between V1, S1, and RL. Incorporating more rigorous spatio-temporal analyses of spontaneous activity could enhance the robustness of these findings.

Here are some important concerns:

(1) The analysis of how spatial topography influences activity correlations in Figure 2 has several issues.

(1a) While squares in V1 and S1 covered a small area of these sensory areas, the correlated territories in RL covered the entire area of RL. The topographic map in V1 continues caudally, so where is the rest of the map in RL? Something similar applies to the relationship between S1 and RL.

We thank the reviewer for pointing out this ambiguity. In the original version, the functional correlation maps were generated using separate seed locations in V1 and S1, and the interpretation relied heavily on thresholded RGB maps in which each pixel was assigned to the color channel with the strongest correlation. This made it difficult to directly compare the V1- and S1-seeded maps and could give the impression that the correlation structure extended differently across RL depending on the chosen seed area.

We have therefore revised the analysis and presentation of Figure 2. Instead of using separate seeds in V1 and S1, we now use common seed locations in RL and compute the correlation of each RL seed with activity across the imaged cortical field. This allows us to ask more directly whether different locations in RL are associated with spatially distinct regions in both V1 and S1. We now show both the raw correlation maps, in which the RGB channels reflect the correlation values for the three RL seeds (new Figure 2E), and the thresholded/maximum channel representation, in which each pixel is assigned to the strongest of the three color channels (new Figure 2F). The raw correlation maps make the correlation structure visible without relying solely on thresholding, whereas the maximum-channel representation highlights the spatial ordering of the strongest correlations.

With this revised analysis, the topographic relationship across V1, RL, and S1 is clearer and no longer depends on comparing separate V1- and S1-seeded maps. We also clarified in the figure legend and Methods how the RGB maps are computed, how the maximum-channel maps are generated, and how thresholded pixels are represented. In addition, we added Supplementary Figure S1 to show further functional-correlation-map examples across PN9, PN10, and PN13 recordings, with seed locations in V1, S1, or RL as indicated in each panel.

(1b) It is essential to know how areas were drawn. High precision is required.

Consistent delineation of cortical areas is absolutely essential for interpreting the functional correlation maps. We have therefore expanded the Methods to describe how cortical areas were delineated from the wide-field recordings. Briefly, recordings were acquired in a field of view defined relative to lambda and the midline, and cortical-area outlines were assigned using published reference maps together with the spatial organization of spontaneous activity patterns and functional correlation maps. This approach follows the procedure we previously validated for developmental wide-field recordings (Leighton et al., 2021).

To make this transparent, we added Supplementary Figure S3, which illustrates how the reference-map-based outlines were overlaid on the imaging field of view and how functional correlation maps and individual network events helped identify the boundaries of V1 and neighboring areas. We also clarified this in the Methods.

(1c) It is not clear if correlated activity means different events in sync or large events that cover 2 or all 3 cortical areas of interest. The figure points to the second option, which contradicts the size of events at these stages, mainly in the oldest mice analyzed here.

The reviewer asks whether the correlations reflect spatially confined events occurring near-synchronously in different cortical areas, or instead large events spanning V1, RL, and S1. To clarify this point, we revised Figure 2 to show representative activity traces and individual frames from the wide-field recordings (new Figure 2B–D). These examples show that activity can be localized to distinct subregions within V1, RL, and S1 while occurring at similar times across areas. Thus, the observed correlations are not well explained by single large events spreading continuously across the entire imaged field.

We have revised the Results and Figure 2 to make this clearer. In addition, the lagged cross-correlation analysis in Figure 2G shows that V1–S1 correlations peak near zero lag and decay for both positive and negative lags, arguing against a stereotyped travelling-wave-like propagation between the two primary sensory cortices as the dominant explanation for the observed correlations.

(1d) It is fundamental to know in detail and provide examples of how the detection of events was performed. For instance, could the dispersion of light from an event in V1 close to RL cause the detection of activity in RL?

The reviewer asks how events were detected in the wide-field recordings and whether light dispersion could lead to false-positive correlations between neighboring areas. We have clarified this point in the Methods. For the functional correlation analyses shown in Figure 2, we did not perform event detection. Instead, the correlation maps were computed from the continuous fluorescence time courses by calculating Pearson correlations between seed region activity and the activity of every pixel in the field of view. Thus, the functional correlation maps do not depend on detecting or assigning individual events.

To address the concern about whether correlations could reflect light spread from large events rather than genuine co-activity across areas, we revised Figure 2 to include representative activity traces and individual frames from the wide-field recordings. These examples show that activity can be spatially confined to distinct subregions in V1, RL, and S1 while occurring at similar times across areas. This argues against the interpretation that the correlations are simply caused by a single event spreading continuously across the imaged field or by light dispersion from one area into another. We have also described the area delineation procedure in more detail in the Methods and added Supplementary Figure S3 to illustrate how activity patterns and functional correlation maps were used to assign outlines of distinct cortical areas.

Although wide-field imaging cannot completely exclude minor contributions from light scattering near area borders, the spatially localized activation patterns and the topographically ordered correlation maps support the interpretation that the correlations reflect genuine nearsynchronous co-activity across V1, RL, and S1.

(2) For the correlations among V1, S1, and RL, it is crucial to have a consistent method to delineate the borders of cortical areas. The authors mention in one sentence that areas were drawn according to a reference map. More details are needed to convince the reader that the borders are accurate, especially because their shape and position change with age.

As described in our response to point 1b, we have expanded the Methods to clarify how cortical-area borders were delineated in the wide-field recordings. Briefly, recordings were acquired in a field of view defined relative to lambda and the midline, and cortical area outlines were assigned using published reference maps together with the spatial organization of spontaneous activity patterns and functional correlation maps. We also added Supplementary Figure S3, which illustrates how the outlines based on reference maps were overlaid on the imaging field of view and how functional correlation maps and individual network events helped identify the boundaries of V1 and neighboring areas. This makes the delineation procedure more transparent across animals and developmental ages.

(3) The results from the model seem to be based on the initial bias in connectivity between neighboring cells from the different areas. Then, it seems straightforward that implementing correlated activity with Hebbian and synaptic depression rules will force the strengthening of connections between spatially close cells. Despite this apparent predisposition of the model towards a defined outcome, the flaws in the experimental data used prevent a rigorous interpretation of the computational model.

We understand the reviewer’s concern that the initial topographic bias could predispose the model toward the emergence of topographic maps. However, the model results show that this bias is not by itself sufficient to determine the final organization (Figure 3B,E,G– I). When V1–S1 correlations are weak, many RL neurons decouple from the primary sensory inputs, resulting in poor topography and few bimodal neurons (Figure 3E,G–I). Conversely, when V1–S1 correlations are very strong, the two input maps become highly aligned, but topography is degraded because many RL neurons receive similar visual and somatosensory inputs at the same topographic location, thereby overriding the initial topographic bias (Figure 3E,G,H). Thus, the initial bias does not simply determine the final map structure. Rather, appropriate topography, map alignment, and the emergence of a mixture of unimodal and bimodal neurons require an intermediate level of correlated activity.

We have revised the manuscript to make this interpretation clearer. We also strengthened the experimental basis for the activity structure used in the model by revising Figure 2 and the corresponding Results and Methods. The revised analyses now show near-synchronous but spatially distinct activation of V1, RL, and S1, a lagged cross-correlation analysis arguing against stereotyped travelling-wave-like propagation between V1 and S1, and functional correlation maps computed from common RL seed locations. Together, these additions clarify the spatial and temporal structure of the spontaneous activity used to motivate the model.

Finally, as described in our response to Reviewer 1, point 4, we have extended the model to test the role of the initial bias more directly in combination with experimentally measured differences in V1 and S1 activity patterns. In this analysis, we incorporated these activity differences and examined how they interact with different levels of initial connectivity bias (Supplementary Figure S2). These simulations show that more mature S1-like activity patterns alone are not sufficient to generate the appropriate topographic and aligned maps, and that an initial connectivity bias is required. At the same time, consistent with Figure 3, this bias does not by itself determine the final organization; the outcome also depends on the temporal correlation structure of V1 and S1 activity.

We agree that the initial topographic bias remains an important modeling assumption, consistent with the idea that coarse activity-independent mechanisms provide an initial scaffold for later activity-dependent refinement. We now present the model accordingly: not as showing that correlated activity alone creates topography from an entirely unstructured circuit, but as showing how structured spontaneous activity can refine an initially coarse topographic scaffold to produce aligned multisensory maps and a mixture of unimodal and bimodal RL neurons.

(4) In the Introduction, the authors nicely and briefly explain the role of primary and higher order sensory cortices in information processing. They also explain how spontaneous activity during development helps to build these circuits by refining connections or establishing hierarchies. They continue explaining the relevance of aligning different topographic maps to allow multisensory integration. Then they provide some examples of sites of multisensory integration. This provides a general context for the data presented in the Results section; however, and importantly, there is no specific introduction of why they are interested in RL and its interaction with V1 and S1. The authors should introduce the RL area and explain why it is an interesting site for multisensory processing.

We thank the reviewer for pointing this out. We have revised the Introduction to make the rationale for focusing on RL more explicit. Specifically, we now introduce RL as a higher-order cortical area located between V1 and S1 that receives topographically organized input from both primary sensory cortices and contains overlapping visual and tactile representations. We also clarify that RL is a particularly relevant area for studying multisensory map alignment because corresponding locations in visual and whisker space can converge onto the same RL neurons, including bimodal neurons. Finally, we expanded the Introduction to explain that RL has been implicated in visually guided tactile behaviors and cross-modal generalization, making it an appropriate model system for studying how aligned multisensory representations emerge during development.

(5) The results shown in Figure 1 corroborate published data from Golshani et al, Rochefort et al, Murakami et al. While the reproduction of data is more than welcome, the authors should specify which part of the data is completely new and acknowledge clearly the rest as corroboration of previous data. The sentence "As described in previous experiments ..." partially acknowledges this fact but is not clear enough. In addition, the transition between this part of the manuscript and the next data is not smooth. Data seems to be used to feed the model so perhaps the organization of the manuscript leaves room for improvement.

We thank the reviewer for pointing this out. We have therefore revised the Results to clarify that the developmental sparsification of spontaneous activity in V1 is consistent with previous work, including Portera-Cailliau, Konnerth, Hanganu-Opatz, Crair and Ohki labs as well as our own (Siegel et al. 2021) and that similar developmental trends in S1 and RL corroborate and extend these observations across the sensory and higher-order cortical areas analyzed here.

We also clarified what is new in the present analysis. Specifically, our contribution is not simply to reproduce previously described developmental sparsification, but to compare V1, S1, and RL within the same experimental and statistical framework, revealing that S1 exhibits more mature activity features earlier than V1 and RL. We also revised the transition to the next section to make clearer how these measurements motivate the subsequent analysis of temporally and spatially correlated spontaneous activity between V1, S1, and RL.

Reviewer #3 (Public review):

Summary:

The study by Dwulet et al. explores how the development of spontaneous neural activity in primary sensory cortices influences the co-alignment of multiple sensory modalities in higher order brain areas (HOAs). To address this question, they focus on connectivity between the primary visual (V1) and somatosensory (S1) cortices and an associative cortical area (RL) in mice. The authors combine experimental (wide-field and two-photon calcium imaging) and computational approaches to show that spontaneous activity matures at a different pace across these brain regions. Their data indicate that S1 develops more rapidly than V1, which is possibly beneficial for RL's integration of visual and somatosensory inputs through correlated spontaneous activity. Using a computational model, they demonstrate that a moderate correlation between V1 and S1 activity can optimally guide the formation of bimodal neurons in RL, which are crucial for maximizing the decodability of multisensory stimuli. This finding highlights the role of correlated spontaneous activity in primary sensory cortices in establishing co-aligned topographic multimodal sensory representations in downstream circuits.

Strengths:

The manuscript is well written and it provides strong enough evidence to support the main claim of the authors. The insights on the role of correlated activity on instructing co-aligned multisensory maps in HOAs are not trivial and are an important advancement for the field.

Weaknesses:

In the opinion of this reviewer, the study has no major weaknesses. A drawback of the work is that none of the predictions of the computational modeling have been corroborated through mechanistic experimental manipulations of early brain activity.

We thank the reviewer for their positive assessment of the manuscript and for highlighting the importance of the model predictions. We agree that a direct mechanistic perturbation of early spontaneous activity would provide an important future test of the model. Such experiments could, for example, perturb the temporal correlation structure between V1 and S1 during the relevant developmental window and then test whether this affects the alignment of V1/S1 maps in RL and the emergence of bimodal RL neurons.

In the present study, we focused on identifying candidate features of spontaneous activity that could instruct multisensory map alignment and testing their sufficiency in a computational model. We now explicitly acknowledge in the Discussion that causal perturbations of early spontaneous activity will be needed to validate the model predictions experimentally. We believe this provides an important direction for future work while preserving the main conclusion of the current study: that structured, moderately correlated spontaneous activity provides a plausible developmental mechanism for refining aligned multisensory representations in higher-order cortex.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Additional comments/suggestions for the figures:

(1) In Figure 1D-G, some of the dots lie almost directly on top of each other, essentially "hiding" certain data points. Using different shapes for each of the three regions might help alleviate this issue and make the data more visually distinct.

We thank the reviewer for this suggestion. We have revised Figure 1D-G so that the three cortical regions are shown with different marker shapes. This should make overlapping data points easier to distinguish and clarify that each point corresponds to the average value for one animal and cortical region at the indicated postnatal age.

(2) In Figure 2D-E, RGB color values are used to represent the highest correlation coefficient across the three seeded areas. It would be more informative if these also depicted the magnitude of the correlations, possibly through a color gradient. Additionally, the black regions in these panels are not currently defined and should be clarified.

We have revised the functional correlation-map analysis and its presentation in Figure 2, as suggested by the reviewer. In the revised figure, the main correlation-map panels now use three seed locations in RL and show the resulting correlations across the imaged cortical field. We present the maps in two complementary ways. First, the raw RGB correlation map shows the correlation values for all three seed locations, with the intensity of each color channel reflecting the magnitude of the corresponding Pearson correlation coefficient (new Fig. 2E). Second, the maximum-channel representation assigns each pixel to the seed location with the strongest correlation, while still preserving correlation strength through pixel intensity (new Fig. 2F).

We have also added color scales to relate pixel intensity to correlation magnitude and clarified that black pixels in the maximum-channel representation correspond to pixels below the correlation threshold used for visualization. The figure legend and Methods now describe how the RGB maps and maximum-channel maps were computed. Finally, we added Supplementary Figure S1 with additional examples from PN9, PN10, and PN13 recordings, with seed locations in V1, S1, or RL as indicated in each panel. This illustrates that they are quite similar across the ages investigated here.

(3) I found Figure 3F a bit difficult to interpret without referring to the Methods section for the definitions of Topography and Alignment. Since these definitions are relatively short and essential for understanding all the modeling figures, I suggest moving them into the main text where they are first introduced.

The definitions of Topography and Alignment have been added to the text where they are introduced.

(4) In Figures 3-5, it is unclear what causes the variability in the model’s responses, as there are two potential sources of randomness: the initial random connectivity matrix and the correlated inputs driving the system. Are either of these fixed? For example, is the distribution of dots along the y-axis in Figure 3G-H, which corresponds to zero correlation between V1 and S1, driven by variability in the initial connectivity matrix, the random timing of input events, or a combination of both? If it’s a combination, it would be interesting to tease this effect apart by fixing one form of randomness and recreating these plots.

In the original simulations in Figures 3–5, neither source of randomness was fixed across runs: each point corresponds to an independent developmental realization with a newly sampled initial connectivity matrix and a newly sampled sequence of spontaneous input events. The initial connectivity was random but weakly biased toward matched topographic location, while spontaneous activity consisted of stochastic independent and shared events activating randomly chosen contiguous groups of neurons (as explained in the main text and Methods). Thus, for example, the spread of points at zero V1–S1 correlation in Figures 3G–H reflects a combination of variability in the initial connectivity and variability in the independent V1 and S1 event histories. At zero correlation, no shared V1–S1 events are present, so this spread does not reflect variability in correlated shared events, but rather run-to-run differences in the two independently refined maps.

We have clarified this point in the text and figure legend. We agree that fixing one source of randomness while varying the other would be an interesting additional analysis to decompose the relative contribution of initial wiring versus input history. However, the goal of the present simulations was to characterize the ensemble of possible developmental outcomes when both initial connectivity and spontaneous activity vary, as expected biologically.

This interpretation is also consistent with the earlier two-layer model from developmental refinements from retina/thalamus to V1 (Wosniack et al., eLife 2021) on which our model builds, where final receptive fields emerge from the interaction between weak biased initial connectivity and stochastic structured spontaneous activity. In the current three-layer extension, the same principle applies to two converging projections, from V1 to RL and from S1 to RL. The initial topographic bias constrains the possible map structure, while the spatiotemporal statistics of V1 and S1 activity determine whether the two maps remain separate, align, or collapse into overly bimodal representations.

(5) The specific parameter values used to create the panels in the modeling figures (Figures 3 and 4) should be made clearer, at least in the figure captions. For example, in Figure 3E, the exact values for the “weak,” “medium,” and “strong” correlations should be provided. Additionally, Figure 4 does not mention the strength of the correlated input considered, which should be specified as well.

The values for the weak, medium and strong correlations have been added to the figure caption of Figure 3. The input correlation for Figure 4 is also now specified in the figure caption.

(6) There is an odd vertical line in Figure 3I that doesn’t appear to be discussed or defined. Its purpose should be clarified, or the line should be removed if it is unintentional.

This line has been removed.

(7) There is a typo in the caption for Figure 3. Panel 'K' should be panel 'J'.

This typo has been corrected.

(8) In the text, the authors write "With these connectivity refinements, the generated activity in RL became sparser in terms of amplitude and participation rate (Figure 3J)." While this appears to be the case for this single example, it is difficult to confirm without zooming in on the panel. These quantities should be computed across multiple instances, and a summary plot should be provided to support this statement.

The experimentally measured developmental sparsification of RL activity is quantified (independent of the model) in Figure 1D–F.

We see how the original wording placed too much weight on the illustrative example in Figure 3J. We have revised the text to clarify that Figure 3J shows a representative simulation illustrating how RL activity changes as V1/S1-to-RL connectivity refines, rather than a separate population-level quantification across model instances.

At the same time, this example is not meant to introduce a new, unsupported mechanism. The model used here is an extension of our previous two-layer model of developmental refinements between retina/thalamus and V1, in which spontaneous activity refined feedforward receptive fields from thalamus to V1. In that study, we specifically quantified how receptive field refinement led to sparsification of cortical activity in V1 over development, including reduced event amplitude, reduced event size/participation, and reduced pairwise correlations (Wosniack et al., 2021). Thus, the example shown in Figure 3J is consistent with a mechanism that has already been systematically characterized in the simpler two-layer setting.

In the present manuscript, the central modeling results concern the emergence of topography, alignment, and the balance of unimodal and bimodal RL neurons. We therefore have softened the corresponding statement and explicitly refer to Figure 3J as an illustrative example.

(9) Figure 5C is a bit difficult to interpret. The corresponding text states, "However, when activity across V1 and S1 is moderately correlated, having some unimodal RL neurons can achieve a higher total maximum fraction of variance for both V1 and S1 compared to the purely bimodal case (Figure 5C)", from which I infer that these dots represent networks resulting from "moderate correlations." However, the exact range of correlations considered should be mentioned in the text or figure caption. Additionally, I find it unusual that some networks with close to 0% bimodal cells perform quite well in reconstructing both S1 and V1. Many data points overlap, but I notice quite a few pale dots in the upper right of the plot. I believe this should be addressed in the main text.

We thank the reviewer for this helpful comment. We have added the correlation values used for the simulations in Figure 5C to the figure caption and clarified the interpretation in the Results. The high reconstruction performance for some networks with relatively few bimodal cells arises because, when V1 and S1 activity are not perfectly correlated, unimodal RL neurons can provide unambiguous information about activity in one sensory area. In contrast, a purely bimodal population can make it more difficult to distinguish whether one or both primary sensory cortices were active. Thus, for moderately correlated inputs, a mixture of unimodal and bimodal RL neurons can reconstruct both sensory areas better than a population composed entirely of bimodal neurons. We have revised the main text to make this point explicit.

(10) The network schematics in Figures 3A and 5A could be improved to better illustrate the network setup using a similar approach as the one used by this research group in Wosniack et al. (2021). Adding arrowheads to the lines from V1/S1 to RL would clarify that these are purely feedforward inputs. It would also be helpful to depict that V1 and S1 are driven by correlated events that are spatially structured.

We thank the reviewer for this helpful suggestion. We have revised the schematics in Figures 3A and 5A to make the feedforward nature of the model clearer by adding arrowheads to the projections from V1 and S1 to RL. We have also clarified the depiction and description of the input activity. Specifically, Figure 3C shows the spontaneous events driving V1 and S1 in the model, including shared events that are both temporally correlated and spatially structured across corresponding topographic locations in the two primary sensory areas. These shared events activate matched contiguous groups of neurons in V1 and S1, while independent events activate randomly chosen contiguous groups within each area. We have clarified this point in the Results and Methods.

General comments regarding the text (including typos):

(1) In Statistical analysis, "In wide-field calcium imaging (we re-analyzed data from [46] (Figure 1))..." should be referencing Figure 2.

Typo fixed.

(2) Right before Table 1, the authors mention that they ran the simulations for 500,000 milliseconds, which is 500 seconds. This doesn't seem long enough for the weights to approach their steady-state values given the inter-event interval. Since the example simulations in Figure 3 are 1,000 seconds long, I'm guessing this is a typo.

Typo fixed. Indeed the simulations in Figure 3 were 1,000 ms (1 s) long.

(3) The specific time step used for the simulations should be specified. Currently, the text only mentions "sufficiently small time steps".

We have now specified the simulation time step in the Methods.

(4) In the Rate-based network model section, you write "These biased weights decay with a Gaussian profile with increasing distance (Figure 3)), with amplitude a and spread s", but Table 1 denotes these parameters differently.

We have corrected the notation so that the parameter names are consistent between the Methods and Table 1.

(5) Currently, all differential equations are written as 1/tau*df/dt. Based on the units of your time constants (seconds), I believe these equations should be tau*df/dt.

We have corrected the differential-equation notation.

(6) Equations 5-6 and 8 should be differential equations.

We have corrected these equations so that they are written as differential equations. These mistakes happened because we changed formats between from Word to Latex.

(7) The expectation in Equation 8 is not clearly defined and I would think here that the W_ij's should be within expectations. In the next paragraph, the authors specify that they are interested in a specific case of W_ij's, but this condition has not been introduced yet.

We thank the reviewer for pointing out this ambiguity. We have revised the text around Equation 8 to define the expectation more clearly and to introduce the specific steady-state connectivity configuration before it is used. Because the expectation is taken over the input activity statistics at steady state, the weights are fixed quantities in this calculation. Including W_ij inside the expectation would therefore not change the result, but we have revised the notation and explanatory text to make this clearer.

(8) The expectation in Equation 8 is not clearly defined, and I believe that the W_ij’s should be included within the expectations (in the following paragraph, the authors mention that they are interested in a specific case of W_ij’s, but this condition has not yet been introduced).

This comment is the same as the one above. Please see the point above for the reply.

(9) At the start of "Optimal weight matrix for correlated input populations", you write that the vector X is M x 1. If that is the case X'X would be a 1x1 matrix. I'm not sure if you meant to write X as 1 x M or to examine XX'.

We thank the reviewer for pointing out this dimensional inconsistency. We have corrected the notation in the Methods. The concatenated input vector X=[v; s] has size M x 1, so the relevant input covariance matrix is X X^T not X^T X. This covariance matrix has size M x M, as required for the eigenvector analysis. We revised the corresponding equations and explanatory text accordingly.

(10) Equation 11 has an s_i on the right-hand side that should be a \mu_s.

Typo fixed.

Reviewer #2 (Recommendations for the authors):

Some sentences may require more scientific rigor. For instance: "We found that activity between the visual and the somatosensory cortex is often, but not always, temporally synchronized.

We have revised the Results to state the quantitative observations more explicitly. Specifically for this example, we now report that the average activity in V1 and S1 across PN9PN12 animals showed a range of Pearson correlation coefficients with a mean of approximately 0.5. We also describe the examples in Figure 2B-D as near-synchronous but spatially distinct activation of subregions in V1, RL, and S1, and we use the lagged cross-correlation analysis in Figure 2G to support the conclusion that V1-S1 correlations peak near zero lag rather than reflecting stereotyped propagation with a fixed delay.

Reviewer #3 (Recommendations for the authors):

Minor suggestions on how to improve some specific aspects of the manuscript.

Introduction:

(1) What do the authors mean when they write "Higher-order areas (HOAs) situated between primary sensory areas"? This sentence might need some editing.

We have revised the sentence to clarify that we are referring to higher-order cortical areas that receive and combine inputs from multiple primary sensory areas. We now also state explicitly that some of these areas, including RL, are anatomically positioned between the primary sensory cortices whose inputs they integrate.

(2) In later portions of the manuscript, it becomes clear what the authors mean when they write “whereby sensory neurons converge onto higher-order cortex while preserving space”, but I think that it would be beneficial if this statement would be better explained also in the introduction.

This has been clarified in the introduction. Specifically, we now clarify that topographic convergence means that neurons representing corresponding regions of sensory space in different primary sensory areas can project to overlapping or nearby locations in higher-order cortex. In the case of RL, this means that visual and tactile representations with corresponding spatial organization can converge onto RL neurons, including bimodal neurons.

(3) Could the authors provide some more information about RL and the rationale as to why it was chosen as the HOA that they investigated in the study?

We have expanded the Introduction to make the rationale for focusing on RL more explicit. We now introduce RL as a higher-order cortical area located between V1 and S1 that receives topographically organized input from both primary sensory cortices. We also explain that RL contains overlapping visual and tactile representations, including bimodal neurons, and that corresponding locations in visual and whisker space can converge in RL. In addition, we now note that RL has been implicated in visually guided tactile behavior and cross-modal generalization. These anatomical and functional properties make RL a particularly suitable model system for studying how aligned multisensory representations emerge.

Results:

(1) "RL was found to slightly lag behind V1 and S1". On what evidence is this statement based upon? As far as I can understand, there are no significant differences between V1 and RL besides amplitudes being higher in RL, which I don't think can be univocally interpreted as a sign that RL lags behind V1 in the developmental profile.

The evidence for a delayed RL maturation relative to V1 and S1 is limited and comes from the pattern of coefficient estimates in the linear mixed models, now shown in Supplementary Tables S1-S3, rather than from a robust difference across all measured activity features. We have therefore revised the Results to state more conservatively that RL and V1 develop more similarly during the second postnatal week, while S1 shows more mature activity features earlier in development. The full linear mixed-model comparisons using V1, S1, and RL as reference areas are provided in Supplementary Tables S1-S3.

(2) Figure 1H is very hard to read.

(a) The slopes and the intercepts have values that differ by orders of magnitude, so the slopes get squeezed and become invisible. Further, the different parts of the plots (e.g. the one of amplitude and duration) are almost overlapping, which is a bit confusing. Slopes and intercepts should also have different units of measure (see Equation 3), so I wonder how they can lie on the same axis. Can the authors try to plot the data in a manner that is easier to visually inspect?

(b) Including the "reference" (V1) intercept in H is also a bit misleading, as one might intuitively interpret it as a difference between V1 and other brain areas. Perhaps the overall differences between brain areas (regardless of age) might be best represented in a plot without age on the x-axis (only brain area). Alternatively, one might point them out directly on the plots in DG.

(c) In D-G, what do the individual dots represent? The legend states N=10 animals, but I only see ~6 dots per plot.

We thank the reviewer for these helpful points. We have revised the caption of Figure 1H and added Supplementary Tables S1-S3, which provide the full linear mixed-model estimates for each choice of reference area. These tables report the intercepts, slopes, interaction terms, confidence intervals, and significance levels in a format that avoids placing quantities with different units and scales on the same visual axis.

For the caption of Figure 1H: The V1 value corresponds to the model intercept at PN8, whereas the age coefficient corresponds to the slope for V1. The S1, RL, Age: S1, and Age: RL terms represent differences relative to this reference model. To avoid the impression that the V1 intercept represents a difference between areas, we now explicitly state that the coefficients in Figure 1H are interpreted relative to V1 at PN8, and that the complete comparisons using S1 and RL as reference areas are provided in Supplementary Tables S2 and S3.

Finally, we clarified that the individual points in Figure 1D–G represent animal-level averages for each cortical area at the indicated age. The value N = 9 refers to the total number of animals included across the dataset, not to the number of animals at each postnatal age. Because recordings were distributed across ages and some points overlap visually, fewer points are visible in individual panels than the total N.

(3) Figure 2B-C: at which lag does this correlation peak? Is it at 0ms? Or does one brain area precede/follow the other one?

We thank the reviewer for this comment. We have revised Figure 2 to include a lagged V1–S1 cross-correlation analysis. The V1–S1 correlation peaks close to zero lag and decreases for both positive and negative lags, indicating that the dominant temporal relationship is near-synchronous rather than consistent with fixed-delay propagation from one primary sensory cortex to the other. The curves show a mild asymmetry, with somewhat stronger correlations when S1 precedes V1, but because the dominant peak is near zero lag, we interpret the data primarily as evidence for near-synchronous, spatially structured coactivity across areas rather than stereotyped travelling-wave propagation. We have added this interpretation to the Results and clarified the temporal-lag convention in the Figure 2 legend.

(4) Figure 2D-E: in the methods section the authors report that "The actual color of each pixel represents the highest coefficient of correlation value across the three channels." I think that this important information should be included in the main text or the legend of the figure.

We have changed Fig. 2 now to clarify the quantification of the functional correlation maps and also added the information requested by the reviewer to the figure legend.

(5) Figure 3D: I think that it would be beneficial if the authors would highlight directly in the figure that those connectivity matrices are between V1/S1 and RL.

This information has been added to the figure.

(6) Figure 3I: does the vertical line correspond to the "critical amount of temporal correlation" (eq. 2)? If so, could the authors provide this information in the figure or the figure legend?

This line was unintentional and has been removed.

(7) It would be nice if the data that was generated for this study (and the data that has already been published and was used to generate Figure 2) would be made publicly available on an open-access repository.

We agree that open data sharing is important. We have made the code used for the model and figure generation available in the repository listed in the Data and Code Availability section. At present, we are not able to deposit the complete raw imaging datasets in an open repository because the wide-field and two-photon imaging files are very large, amounting to multiple terabytes, and we do not currently have a sustainable hosting solution for these raw data. We will share data upon request, and we will deposit the raw imaging datasets in an appropriate open repository if a feasible long-term hosting solution becomes available.

References

M. Chini, T. Pfeffer, and I. Hanganu-Opatz. An increase of inhibition drives the developmental decorrelation of neural activity. eLife, 11:e78811, 2022.

P. Golshani, J. T. Gonçalves, S. Khoshkhoo, R. Mostany, S. Smirnakis, and C. PorteraCailliau. Internally mediated developmental desynchronization of neocortical network activity. Journal of Neuroscience, 29(35):10890–10899, 2009.

A. Gribizis, X. Ge, T. L. Daigle, J. B. Ackman, H. Zeng, D. Lee, and M. C. Crair. Visual cortex gains independence from peripheral drive before eye opening. Neuron, 104(4):711–723.e3, 2019.

S. Lakhera, E. Herbert, and J. Gjorgjieva. Modeling the emergence of circuit organization and function during development. Cold Spring Harbor Perspectives in Biology, 17(2):a041511, 2025.

A. H. Leighton, J. E. Cheyne, G. J. Houwen, P. P. Maldonado, F. De Winter, C. N. Levelt, and C. Lohmann. Somatostatin interneurons restrict cell recruitment to retinally driven spontaneous activity in the developing cortex. Cell Reports, 36(1):109316, 2021.

H. Matsumoto, T. Murakami, and K. Ohki. Topographic correspondence between retinotopic and whisker somatosensory map in mouse higher visual area and its development. Frontiers in Neural Circuits, 19:1552130, 2025.

T. Murakami, T. Matsui, M. Uemura, and K. Ohki. Modular strategy for development of the hierarchical visual network in mice. Nature, 608:578–585, 2022.

N. L. Rochefort, O. Garaschuk, R.-I. Milos, M. Narushima, N. Marandi, B. Pichler, Y. Kovalchuk, and A. Konnerth. Sparsification of neuronal activity in the visual cortex at eyeopening. Proceedings of the National Academy of Sciences of the United States of America, 106(35):15049–15054, 2009.

Version 2

Abstract

To perceive and navigate complex sensory environments, animals combine sensory information from multiple modalities in specialized brain circuits. Known as multisensory integration, this process typically depends on the existence of co-aligned topographic connections from several sensory areas to downstream circuits exhibiting multimodal representations. How such topographically co-aligned connectivity necessary for multisensory integration gets set up in early stages of development is still unknown. Inspired by the role of spontaneous activity in refining topographic connectivity between early sensory circuits, here we investigated the potential of such spontaneous activity to also guide the co-alignment of multiple sensory modalities in RL, a higher-order associative cortical area rostro-lateral to V1. Analyzing spontaneous activity simultaneously recorded in primary visual and somatosensory cortex and area RL at different developmental ages before sensory experience, we identify candidate features of this activity to guide the emergence of co-aligned topographic multisensory projections with somatosensory leading the visual projection. We confirm this hypothesis using a computational model of activity-dependent circuit refinement, and show that the correlation of spontaneous activity between the visual and somatosensory primary cortex can establish an optimal fraction of multisensory neurons in RL for stimulus decoding. Our model provides an exciting new computational perspective of the role of spontaneous activity in the emergence of topographically co-aligned multimodal sensory representations in downstream circuits, specialized for the processing of rich sensory environments.

Introduction

Sensory environments are complex – rich with objects and events that specific brain circuits must decipher to ensure their accurate perception and to guide appropriate behavior. To efficiently perceive and navigate sensory environments, these brain circuits combine information from multiple sensory modalities, including vision, somatosensation, or hearing, a process known as multisensory integration. Several brain regions have been shown to generate multimodal representations, typically by aligning connections from the sensory periphery where specific sensory features are first encoded to downstream projection areas. In the rodent, whiskers transmit tactile sensations which are combined with visual information to guide many behaviors including goal-directed behavior [1], gap crossing and food source detection [2], navigation [3], prey-capture behavior [4], and recognizing object features and shapes [5]. Multisensory integration has been reported in subcortical areas such as the superior colliculus [6, 7] and the striatum [8]. Higher-order areas (HOAs) also play play a critical role in multisensory integration, because they receive and combine inputs from multiple primary sensory areas [1, 3, 9–11]. Some of these higher-order areas, including RL, are anatomically positioned between the primary sensory cortices whose inputs they integrate. Moreover, connectivity is organized in a topographic manner across the modalities, whereby neurons representing neighboring locations in sensory space project to neighboring locations in higher-order cortex [3, 12]. For multisensory integration, this means that visual and somatosensory inputs representing corresponding regions of space can converge onto overlapping or nearby locations in higher-order cortex. Specifically, visual and tactile stimuli have overlapping representations in a higher order association area, area RL, that lies between the primary visual (V1) and the primary somatosensory (S1) cortices and receives topographically organized input from both areas. There is significant spatial alignment between V1 and S1 in area RL; e.g. the lower whisker pad aligns to the lower visual field of view and can elicit responses in the same neurons, called bimodal neurons, in RL [3]. RL is therefore a particularly useful area for studying how topographically aligned visual and somatosensory inputs are integrated in higher-order cortex and how such aligned multisensory representations emerge during development. Additionally, this multisensory integration allows for cross-modal generalization. For example, rodents can adapt to changing sensory environments and can generalize sensorimotor tasks to a different modality, i.e. visual, after learning a task from only tactile stimulus, due to aligned abstract representations of these modalities in RL [1], further supporting RL as a relevant model system for studying multi-sensory integration. Finally, evidence shows that rats combine and integrate visual and tactile signals in a supralinear manner, significantly enhancing sensory processing efficiency [5]. This integration is supported by a substantial population of bimodal neurons in RL, which respond to both visual and tactile inputs. Although full functional maturation of visual HOAs occurs only after eye opening, basic properties such as responsiveness to stimuli and topographic organization between V1 (as well as S1) and RL are already present at eye opening [13, 14]. These anatomical and functional properties make RL an ideal model system for asking how aligned multisensory representations are established before sensory experience.

Before the onset of sensory experience, primary sensory cortices generate spontaneous activity, which plays a pivotal role in their proper maturation in terms of cellular properties and activity-dependent connectivity refinements [15–18]. Across the sensory areas, a general developmental profile of spontaneous activity can be observed: early global and dense spontaneous events transition to more localized and sparse patterns, typical of adult networks [15, 17, 19–21]. By promoting the refinement of appropriate connectivity [22–24], these spontaneous activity patterns prepare the developing cortex for future sensory input [25–27], but at the same time are affected by changing circuitry [28]. Spontaneous activity has been characterized and shown to guide the establishment of hierarchical architectures in HOAs in the visual system before eye opening [29], but much less is known about its role in the development and topographic co-alignment of multiple sensory modalities in multisensory HOAs.

Whereas activity-independent mechanisms are known to set up the first cues instructing axons where to go [30, 31], activity-dependent synaptic plasticity mechanisms instructed by either spontaneous activity or sensory experience, are crucial in refining this imprecise connectivity [22, 32, 33]. Indeed, various forms of Hebbian synaptic plasticity, which promotes the potentiation of synaptic strength as a result of correlated pre- and postsynaptic activity, have been proposed to guide the activity-dependent refinement of circuit connectivity in development [28, 34–38]. Some experimental work has also measured Hebbian-like learning rules specific to development [39]. Activity-dependent mechanisms likely also contribute to the establishment and alignment of maps from different sensory modalities to downstream circuits, which has been mainly investigated in the rodent superior colliculus and avian optic tectum [32]. For example, how visual experience aligns the auditory and visual maps during development has been mainly studied by manipulations that cause the two maps to misalign [40–44]. This suggests that these same principles might apply not only to realign and maintain maps after manipulation, but also to the activity-dependent emergence of aligned representations in developing circuits.

Here, we combine computational modeling and calcium imaging experiments to investigate the role of spontaneous activity in the emergence of aligned topographic maps from two primary sensory cortices (visual, V1 and somatosensory, S1) to RL, an associative cortical area rostro-lateral to V1, and the emergence of bimodal responses in RL. We find that spontaneous activity in V1, S1, and RL sparsifies during development, but S1 exhibits more mature activity patterns compared to V1, which desynchronize as early as postnatal day eight. Using these experimental insights, we expand a computational model of activity-dependent connectivity refinement between primary sensory areas (V1 and S1) and RL using biologically plausible Hebbian synaptic plasticity which was previously used to study topographic map-formation and receptive field refinement between the thalamus and V1 [28]. Our model predicts that aligned topographic connectivity refinement and the emergence of bimodal neurons requires an intermediate amount of correlation between the primary sensory cortices (V1 and S1), which we investigate in our data. This connectivity refinement through spontaneous activity between the primary sensory areas and RL is sufficient to explain the developmental sparsification in RL whereby RL neurons become responsive to either or both primary sensory cortices in a ratio that ensures optimal information processing.

Results

Spontaneous activity in V1, S1 and RL during development

To investigate how spontaneous activity matures differentially across the mouse’s developing primary visual (V1), somatosensory (S1), and higher-order areas, specifically area RL, we performed two-photon imaging with one field of view per developing cortex in unanesthetized mouse pups at postnatal days (PN) 8-16 (Figure 1A,B). Spontaneous activity was observed across all three cortical areas (Figure 1C). We used a linear mixed model analysis [45] to estimate the differential effect of age on different aspects of spontaneous activity. Consistent with previous studies of developing visual cortex [15, 17, 21, 46], we found that spontaneous activity in V1 sparsifies throughout development, i.e. the average amplitude, duration, and participation rate of spontaneous activity decrease with age, as reflected in the significant negative effect of age on these properties (Figure 1D-F, H, see Methods). In contrast, the frequency of events increases with age, as reflected in the significant positive relationship between age and event rate (Figure 1G, H). Similar trends of activity sparsification were also observed in S1 and RL. Thus, while the developmental sparsification of spontaneous activity corroborates previous work in primary sensory cortex, our analysis extends these observations by comparing V1, S1, and RL within the same experimental and statistical framework. There were no differences between participation rate across areas, with all areas following a similar decreasing trend over time (Figure 1F,H). Furthermore, comparing area RL to V1 also showed no significant differences in duration of events and event rate between these two areas (Figure 1E,G,H), although amplitudes were slightly higher in RL compared with V1. However, we observed some significant differences, especially in S1. Specifically, spontaneous events in S1 have significantly lower amplitudes, lower durations, and higher rates compared to the other regions, at younger ages (Figure 1D,E,G,H).

Spontaneous activity in RL sparsifies over development, with more mature activity in S1 compared to V1 and RL.

A. Primary visual cortex (V1), somatosensory cortex (S1), and rostro-lateral area (RL) imaged in sequence over a 100 min recording session. B. Sample field of views of the individual cortices. Scale bar corresponds to 50 µm. C. Sample z-profile traces from the three different cortices (columns) in three mice at different postnatal ages (rows). Vertical scale bar corresponds to 20% ΔF/F0. Horizontal scale bar corresponds to 50 seconds. D-G. Amplitude (D), duration (E), participation rate (F), and event rate (G) as a function of postnatal age. Each dot represents the average value for one animal for one cortical area at the indicated postnatal age, computed across all detected events in that area. N = 9 was the total number of animals included across all ages. Because animals were distributed across postnatal ages and some points overlap visually, fewer than 9 points are visible at any individual age. Regression lines derived from the linear mixed model in H. H. Linear mixed-model estimates for each activity feature, where age and region are included as fixed effects, and mouse ID and sample ID are included as random effects. Values are shown relative to V1 at PN8. The V1 term corresponds to the model intercept, the Age term corresponds to the age-dependent slope in V1, and the S1, RL, Age:S1, and Age:RL terms indicate differences from this reference (see Methods). Stars indicate statistical significance (* p<0.05, ** p<0.01, *** p<0.001). Because intercepts and slopes have different units and scales, full model estimates, confidence intervals, and comparisons using S1 or RL as reference areas are provided in Supplementary Tables S1-S3.

Together, this suggests that spontaneous activity in S1 has more mature amplitudes, duration, and event rates earlier in development, whereas RL and V1 develop more similarly during the second postnatal week. In summary, spontaneous activity in developing primary cortices and RL sparsifies with different temporal profiles, with the somatosensory cortex showing earlier maturation than the visual cortex and RL (Supplementary Tables S1-S3).

The primary somatosensory and visual cortex exhibit correlated spontaneous activity

Having established that spontaneous activity matures with different temporal profiles across V1, S1, and RL, we next asked whether activity in V1 and S1 is temporally and spatially correlated in a way that could support the emergence of aligned multisensory representations in RL. To this end, we reanalyzed our published wide-field calcium imaging recordings from both V1 and S1 [47] to assess the temporal relationship between spontaneously occurring activity in these two primary sensory cortices (Figure 2A). We found that activity between the visual and the somatosensory cortex is often, but not always, temporally synchronized, with examples showing near-synchronous but spatially distinct activation of subregions in V1, RL, and S1 (Figure 2B-D). Indeed, analyzing the average activity between V1 and S1 of several animals at postnatal days PN9-PN12 revealed a range of Pearson correlation coefficients, with a mean of ~0.5 (Figure 2H).

Spontaneous activity in V1, S1, and RL is near-synchronous and spatially organized.

A. Schematic of the wide-field experiments. Cortical surface area encompassing the primary visual (V1) and somatosensory (S1) cortices as well as RL in a PN9 mouse. Higher-order areas of the visual cortex are drawn according to [9]. LM: lateromedial area, AL: anterolateral area, A: anterior area. B. ΔF/F0 traces from 3 ROIs in V1, S1 and RL (shown in A). C. Expanded view of the traces shown in the area marked by a light grey box in B. D. Four ΔF/F0 frames over a period of 900 ms indicated by a dark grey vertical bar in C. E. Functional correlation map of the recording shown in B-D. Pixel values represent Pearson correlation coefficients between each pixel and the areas marked by blue, green and red squares in RL across the recording’s duration. The color of each pixel is the RGB composite of the coefficients of the areas marked by the three colored squares in RL. F. Same as E except that each pixel is assigned to the color channel (red, green or blue) with the highest coefficient of correlation value across the three channels. G. Pearson correlation between activity in somatosensory and visual cortex as a function of temporal lag at PN9. Individual animals are shown in gray and the mean in black. The vertical black line indicates zero lag. Arrows indicate the temporal order of activity at different lags. H. Temporal correlations between V1 and S1. Violin plots where each dot indicates the Pearson correlation coefficient for one 2-minute bin. PN9 N = 6; PN10 N = 5; PN11 N = 4; PN12 N = 2.

To test whether these correlations reflected a consistent propagation of activity from one primary sensory cortex to the other, we computed V1–S1 correlations at PN9 as a function of temporal lag. Correlations were maximal near zero lag and decreased for both positive and negative lags, arguing against stereotyped traveling-wave-like propagation between V1 and S1 (Figure 2G). The lagged correlation curves showed a mild asymmetry, with somewhat stronger correlations when S1 preceded V1. However, because the dominant peak was near zero lag, we interpret these data primarily as evidence for near-synchronous coactivity rather than a consistent temporal lead of one area over the other. Together with the examples of near-synchronous but spatially distinct activation in V1, RL, and S1, this suggests that spontaneous activity across these areas is predominantly near-synchronous and spatially structured, rather than reflecting a single wave propagating across the entire imaged region.

Previous work has shown that connectivity between V1 and RL as well as S1 and RL is topographically organized in mature animals to preserve space [3, 9]. Furthermore, this organization is also aligned across the modalities, i.e. stimuli in the same region of sensory space tend to converge onto the same cells in

RL [3, 12, 29]. Hence, in addition to temporal correlations, we evaluated whether spontaneous activity also carries information about spatial topography between the two primary sensory cortices and RL. We computed functional correlation maps for three distinct seed locations in RL, showing that spontaneous activity in RL is coordinated topographically with activity in both V1 and S1 (Figure 2E,F). Additional functional-correlation-map examples from PN9, PN10, and PN13 recordings showed similar spatially ordered correlation structure across the developmental window analyzed here, with seed locations placed in V1, S1, or RL as indicated in each panel (Supplementary Figure S1). This is consistent with previous work showing that functional-connectivity maps computed from spontaneous activity can reveal topographic organization in higher visual areas before eye opening [29], and with recent evidence that retinotopy-like and somatotopy-like patterns of ongoing activity, together with their rough topographic correspondence in RL, are already present at P10–11 [48]. Thus, spontaneous activity between V1 and RL and between S1 and RL is also spatially correlated and carries topographic information. This suggests that some activity-dependent mechanism might be using these temporal and spatial correlations during development to establish topographic connectivity maps between the primary sensory areas and RL, and hence enable multisensory integration in the adult animal.

Correlated spontaneous activity can establish topographic connectivity between the primary sensory cortex and RL and generate bimodal cells in RL

To test the hypothesis that an activity-dependent mechanism may use the spatiotemporally structured spontaneous activity in the primary sensory cortices to set up topographic connectivity from each area into RL and generate bimodal cells, we built and analyzed a computational model with three populations of neurons for V1, S1, and RL (Figure 3A) [28]. We modeled the activity-dependent synaptic plasticity of feedforward connections between V1 and RL, and S1 and RL, with each area subject to spontaneous activity as characterized before (Figure 1). While we initialized the connectivity as random, we included a slight bias for neighboring cells in the primary sensory cortices to project onto neighboring cells in RL (Figure 3B, Methods). This bias incorporates activity-independent effects like molecular gradients that are known to set up a coarse topography before the onset of activity-dependent plasticity, at least for connectivity between the sensory periphery and subcortical or primary cortical areas [38, 49–51]. We therefore interpret this bias as a coarse initial scaffold on which activity-dependent plasticity acts, rather than as a mechanism that by itself determines the final topographic organization. Activity-dependent plasticity was implemented through a synaptic plasticity rule involving Hebbian and heterosynaptic terms. The Hebbian term potentiates synaptic weights based on coincident pre- and postsynaptic activity. The additional heterosynaptic term was implemented to depress synaptic weights, but only in the presence of presynaptic activity (see Methods) (1) change in connection strength=pre × (post - offset).

Model set-up and assumptions.

A. Schematic of a three-population feedforward model of the developing V1, S1, and RL. B. Example of random initial connectivity matrix between V1 and RL (connectivity matrix for S1 and RL is chosen identically) with a topographic bias denoted by stronger connections along the diagonal inspired by activity-independent mechanisms (see Methods). C. Sample of spontaneous events in V1 (top) and S1 (bottom) across cells as a function of time. Boxes highlight two spatiotemporally correlated events in V1 and S1 (see Methods). D. Example of refinement of connectivity matrix between V1 and RL (top) and S1 and RL (bottom), from t0 to tN. Same axes as in Figure 2B. E. Example of connectivity matrices in steady state from simulations with low (correlation=0.1, left), medium (correlation=0.5, center) to highly (correlation=0.9, right) temporally correlated activity between V1 and S1. F. Schematic to illustrate topography and alignment measures (see Methods). G-I. Topography (G), alignment (H), and percentage of bimodal cells (I) as a function of V1 to S1 correlation. Red dots correspond to the examples from E. Each point represents an independent simulation run in which both the initial connectivity matrix and the stochastic sequence of spontaneous events were resampled to mimic different developmental realizations. J. Excerpt of activity in RL at the beginning (left) and end (right) of the simulation (corresponds to panel D, left and right).

This phenomenological plasticity rule has a biophysical basis and was previously derived from a set of interacting molecules (in particular, neurotrophins) [39, 52, 53] and shown to produce selectivity to correlated inputs from random initial connectivity [37, 54]. We then implemented spatiotemporally correlated spontaneous activity in areas V1 and S1, as measured experimentally, which this synaptic plasticity rule used to modify the synaptic weights in the network (Figure 3C). Specifically, S1 and V1 neurons were driven by a combination of independent events within each area and shared events across the two areas. Independent events activated randomly chosen contiguous groups of neurons within either V1 or S1, implementing spatial correlations within each area, which we have previously shown can establish topographic connectivity between two areas [28]. Shared events implemented spatiotemporally structured correlations between V1 and S1: they occurred simultaneously in both areas and activated the same contiguous group of neurons in terms of topographic identity. The common inputs giving rise to shared events might reflect shared inputs from the subcortical areas such as the thalamus or superior colliculus into the primary sensory cortices [55–57] or correlation induced by direct connections between V1 and S1 [58, 59].

Using this spatiotemporally correlated spontaneous activity in the primary sensory cortices, the Hebbian plasticity rule (Eq. 1) refined and stabilized connectivity between V1/S1 and RL (Figure 3D), as shown before in a two-layer network [28]. We investigated the emergent connectivity properties for a range of V1 and S1 temporal correlations (Figure 3E). To quantify the topographic organization of connectivity between V1 and RL, and S1 and RL, and the alignment of maps from V1/S1 to RL, we defined two mathematical measures, topography and alignment (Figure 3F, see Methods). Topography was defined as how well the initial topographic bias was preserved in the final receptive field between either V1/RL or S1/RL and was calculated as the correlation between the final receptive field of each sensory area with RL and an ideal receptive field (see Methods). Alignment was defined as the measure of similarity between the final receptive fields of V1 and S1 and was calculated as the correlation between them. Weak temporal correlations (where independent events dominate over shared events) resulted in strong competition between the primary sensory cortices with dominant synaptic weight depression. The strong competition generated sparse connectivity matrices to RL, with many RL neurons decoupling from the primary sensory cortices (Figure 3E, left). This resulted in poor topography and map alignment of the maps between V1/S1 and RL (Figure 3G,H). RL had almost no bimodal cells, i.e. neurons that receive input from both primary sensory cortices, in contrast to what is found in mature RL experimentally [3] (Figure 3I). Increasing temporal correlations between the primary sensory cortices increased alignment of the V1/S1 to RL maps (Figure 3H). However, very strong correlations were detrimental to V1/RL and S1/RL topography (Figure 3E,G). Strong temporal correlation between V1 and S1 rendered many RL neurons selective to visual and somatosensory inputs at the same topographic location as reflected in the identical connectivity matrices (Figure 3E, right), overriding the initial topographic bias (Figure 3B). Only a medium-strength temporal correlation between V1 and S1, as found experimentally (Figure 2C), generated a connectivity profile which prevented RL cells from decoupling from V1 and S1, while preserving topography (Figure 3E, middle).

We quantified the percentage of bimodal neurons in our model as a function of the temporal correlation between V1 and S1 (Figure 3I). By performing a steady-state analysis of our model (see Methods), we derived the critical amount of temporal correlation, c∗, between S1 and V1 required for the emergence of bimodal neurons. This parameter depends solely on two parameters of the model, i.e. the average firing rate of cells in the sensory cortices, ⟨pre⟩, and the heterosynaptic offset from Eq. 1, (2) c∗=offset−⟨pre⟩. Together, these results demonstrate that the initial topographic bias was not sufficient to determine the final map structure. Instead, appropriate topography, alignment, and bimodality emerged only for an intermediate level of V1-S1 correlation (Figure 3B,E,G-I).

In a representative simulation, the refinement of V1/S1-to-RL connectivity was accompanied by sparser RL activity, visible as reduced event amplitude and participation rate over the course of learning (Figure 3J). This was consistent with the sparsification observed in the experimental data (Figure 1C,D,F) as well as with the activity sparsification previously quantified in the two-layer model of developmental refinements [28].

In summary, our proposed computational model shows how correlated spontaneous activity between V1 and S1 can refine initially biased feedforward connectivity to RL. The model predicts that only medium-strength correlations between primary cortices produce topographically organized and aligned connectivity in V1/RL and S1/RL, and a mixture of bimodal and unimodal neurons in RL consistent with experimental data.

More mature S1 to RL connectivity can contribute to map alignment between V1 and RL

Having established that the spatiotemporal correlations between V1 and S1 can set up topographic and aligned connectivity to RL as well as generate bimodal cells in RL, we next turned to investigate a possible role of the differentially more mature spontaneous activity in S1 compared to V1 (Figure 1H). To examine this, we systematically varied the bias of the topographic connectivity between S1 and RL (Figure 4A). While a stronger bias improved the topographic connectivity between S1 and RL (Figure 4B, right), it also improved the topographic organization of connectivity between V1 and RL (Figure 4B, left, and C) as well as the alignment of connectivity from V1 and S1 to RL (Figure 4D) and affected the number of bimodal cells in RL (Figure 4E). Significantly reducing the connectivity bias from S1 to RL while using more mature S1 activity properties, including lower event amplitude and higher event frequency as observed experimentally (Figure 1D-G), quickly deteriorated topography and alignment (Supplementary Figure S2). Thus, more mature S1-like activity patterns alone were not sufficient to generate appropriately topographic and aligned maps. Rather, an initial S1-to-RL connectivity bias was required, together with the appropriate level and structure of correlated V1-S1 activity described above (Figure 3B,E,G-I).Therefore, a more mature S1 to RL connectivity bias, together with structured spontaneous activity, may contribute to the refinement of connectivity between V1 and RL.

More mature spontaneous events in S1 effectively instruct map alignment between V1 and RL.

A. Three pairs of connectivity matrices (V1 left, S1 right) with increasing (top to bottom) S1 bias. B. Corresponding connectivity matrices in steady-state after plasticity. C-E. For all simulations, V1-S1 correlation of 0.5 was used. C S1 and V1 to RL topography. Different S1-to-RL connectivity biases indicated by color. D. Alignment of V1 and S1 to RL topography as a function of S1-to-RL connectivity bias. Linear fit with R = 0.44, p value< 0.001. E. Percentage of bimodal neurons in RL as a function of S1-to-RL connectivity bias. Linear fit with R = 0.42, p value< 0.001.

mixture of bimodal and unimodal neurons in RL performs optimal decoding of activity

RL is involved in integrating information from different sensory modalities to obtain a more accurate representation of statistically rich sensory environments. In particular, an area that integrates visual and tactile stimuli is able to more accurately infer such information; e.g. when rats use both tactile and visual information, their performance in a task involving determining object orientation is better than the sum of the performance of each modality individually [5]. To investigate how the topographically aligned cortical connectivity between V1 and RL, and S1 and RL, might facilitate the integration of information from different sensory modalities, we quantified how much information about the activity of each primary sensory cortex can be decoded from the activity in RL. We quantified the amount of decoded information as the coefficient of determination, R2, which can be interpreted as the fraction of variance in V1 or S1 that can be reconstructed from activity in RL. To do this, we fitted a linear regression model using RL activity and both V1 and S1 activity, and calculated R2 from the error between our fitted linear model and each sensory area individually, which we refer to as the ‘V1 or S1 reconstruction’ (Figure 5A). [60, 61] (See Methods). R2 is naturally affected by how much activity in V1 or S1 is propagated into RL. In our model (Figure 3) and in biology [9, 62], some regions of the primary sensory cortices are not connected to RL; for example, RL contains only a partial representation of the visual field, hence V1 cells outside of that region do not project to RL [11, 63]. Consistently, we found that the fraction of variance reconstructed, R2, for either V1 or S1 independently increased with the percentage of connected neurons from the respective cortex into RL (Figure 5B, see Methods).

A mixture of bimodal and unimodal neurons can optimally decode activity.

A. Schematic of the regression procedure to calculate R2 for reconstructing V1 or S1 activity from RL activity with a non-linear transfer function (a sigmoid) (see Methods). A linear regression is performed using the non-linearly transformed RL activity (dependent variable) and both V1 and S1 activity (independent variables). R2 is calculated from V1 or S1 activity separately using the fitted regression line. B. Reconstruction of V1 or S1 activity (R2) as a function of the percentage of connected cells from the respective cortex over all values of correlation chosen uniformly between 0 and 1 (see Methods). Color indicates topography. C. Reconstruction of V1 and S1 activity (R2) as a function of the percentage of bimodal cells (denoted by color). For all simulations, V1-S1 correlation of 0.5 was used.

We showed analytically that a locally connected network, where every primary sensory cortical neuron makes at least one connection to RL and every RL neuron receives at least one connection from the primary sensory cortices, can maximize the total fraction of variance reconstructed for both V1 and S1 if the activity in V1 or S1 is perfectly temporally correlated (correlation value of 1) (see Methods). These conditions correspond to a network in which all neurons in RL are bimodal. However, when activity across V1 and S1 is moderately correlated (correlation value of 0.5), this fully bimodal solution is no longer necessarily optimal. For simulations with V1–S1 moderate correlation, networks containing a mixture of unimodal and bimodal RL neurons often achieved higher joint reconstruction of V1 and S1 activity than networks composed almost entirely of bimodal neurons (Figure 5C). This occurs because unimodal RL neurons provide unambiguous information about activity in one sensory area when the two inputs are not perfectly correlated, whereas purely bimodal neurons make it more difficult to distinguish whether one or both sensory cortices were active. Thus, for moderately correlated inputs, unimodal neurons can complement bimodal neurons and improve reconstruction of both sensory areas. This suggests that the mixture of unimodal and bimodal neurons in RL [3] is the optimal solution for encoding the maximum amount of information from the activity in V1 and S1 [64].

Discussion

We investigated the role of spontaneous activity in the maturation of connectivity from two primary sensory cortices, V1 and S1, to RL, a higher-order association area, by combining computational modeling with experiments. We found that while spontaneous activity sparsifies during development in all investigated areas, S1 exhibits relatively mature, desynchronized activity patterns as early as postnatal day eight. Using these experimental insights, we constructed a computational model of activity-dependent connectivity refinement from each V1 and S1 to RL. To generate topographic connectivity from V1 and S1 to RL, aligned connectivity maps for the two sensory projections and a mixture of unimodal and bimodal neurons in RL, our model predicts that spontaneous activity between V1 and S1 should exhibit moderate temporal correlations, which we also found experimentally. Furthermore, we demonstrated how a more mature somatosensory cortex can guide the refinement of visual cortex connectivity to RL, suggesting that inter-cortical influence plays a central role in the establishment of refined and topographically aligned connectivity during development. Finally, we found that moderate correlations between V1 and S1 result in a mixture of bimodal and unimodal neurons in RL which can optimize information processing in RL when either one modality or both are active. Thus, by combining modeling with experiments, we found that activity-dependent plasticity can instruct the formation of refined and topographically aligned connectivity between primary sensory and high-order cortices in development to produce efficient information integration from multiple modalities.

Origin of correlated events in the primary sensory cortices

We find that early spontaneous activity with medium correlations can lead to the optimal amount of bimodal neurons and alignment of these sensory maps before sensory experience, which is crucial for effective multisensory integration in higher-order cortical areas. Such medium correlations indeed exist in the experimental data (Figure 2C). This result makes the prediction that alterations in the correlations in early spontaneous activity might affect later multisensory integration, which can be empirically tested. For instance, spontaneous activity in V1 has been shown to have different correlation properties in a mouse model of Fragile X [65] and in cases of disrupted inhibition [19, 47]. Our model therefore suggests that multisensory integration in RL may be perturbed in these cases. A key future test will be to perturb the temporal correlation structure of early V1/S1 spontaneous activity directly and determine whether this alters the alignment of V1/S1 maps in RL and the emergence of bimodal RL neurons.

What might be the origin of the correlated activity we observed in the visual and somatosensory cortex (Figure 2)? One possibility is that the source is external to the animal [66], i.e. a single stimulus with both visual and somatosensory aspects. Indeed, while a substantial portion of activity in the developing cortex occurs independently of sensory stimulation [23], both the immature visual [67] and somatosensory [68] cortices exhibit some degree of sensory-driven activity. Although it is unlikely that this external stimuli gives any topographic information before eye opening, it could help to correlate the activity between sensory cortices. Alternatively, subcortical areas like the (primary or higher-order) thalamus are in an ideal position for coordinating activation across distinct regions of the cortex [29, 56, 69]. The developing thalamus exhibits spontaneous propagating waves, which have been shown to guide the formation of sensory maps to the cortex and support circuit plasticity [56,70,71]. Spontaneous calcium waves propagate among the different sensory-modality thalamic nuclei up to the cortex and hence provide a means of communication among sensory systems [71]. Finally, correlated events in V1 and S1 could also originate from lateral long-range connections between the visual and somatosensory cortex. These types of connections emerge during the first two postnatal weeks in mice [72] and potentially directly propagate activation from one area to another [58, 59]. More specifically, it has been shown that postsynaptic excitatory neurons in a subregion near the anterior border of V1, situated in close proximity to S1, receive input from local fast spiking inhibitory neurons that have direct cortico-cortical input from the posterior area of S1 [73]. This cross modal suppression may contribute to these medium level correlations during development.

Precise timing of map formation

We have shown that the somatosensory cortex’s early maturation is able to drive the aligned topographic development of the visual cortex Figure 4. This agrees with previous work showing earlier development of somatosensory cortex. For example, layer 4 (L4) to L2/3 excitatory synapses in somatosensory cortex were shown to mature 2 weeks earlier than those in visual cortex [74]. A similar template for the maturation of one map provided by another is observed in the refinement of cortico-collicular connections. Projections from S1 to the superior colliculus arrive as early as postnatal day 2 while projections from V1 enter the colliculus around postnatal day 6 [75], giving more time for S1 maps to refine. The precise timing of connectivity refinement between S1 and the superior colliculus likely guides the later developmental connectivity refinement between V1 and the colliculus and helps align the two maps [75, 76]. Moreover, this process of visual and somatosensory map alignment in the superior colliculus seems to be activity-dependent, with spontaneous activity in the retina playing an important role by driving the alignment of cortico-collicular connections from V1 to a genetically-induced duplicated retino-collicular map [76]. This modality difference in map refinement with the somatosensory leading the visual map seems to be present also in the refinement of connections between S1 vs. V1 and RL as we have shown earlier. Although the activity-dependent plasticity mechanisms behind this are experimentally unknown, inspired by earlier works [28, 52], our models suggest that similar activity-dependent plasticity involving Hebbian and heterosynaptic terms are sufficient.

Role of recurrent connectivity

Neural dynamics in the adult cortex are strongly affected by recurrent connections [77, 78]. In our model, we have chosen to focus on feedforward connections between the primary sensory and RL for several reasons. First, recent experimental evidence from the developing somatosensory cortex suggests that feedforward connectivity might be established first [79], and second, the strength and functional specificity of recurrent connections in the vsual cortex increases only later in development [80]. Furthermore, abstraction is important for deriving robust biological insights from models [81] and here we are focused on how these feedforward connections arise and refine themselves over development building on our previous work of refinements between the thalamus and V1 [28].

Uni- and bimodal neurons in other species

Neurons responding to only a single or to multiple modalities are common in other animals apart from mice, like macaques [82–84], cats [85–87], or ferrets [88, 89]. The surprising presence of unimodal neurons in association cortices has inspired a number of studies attempting to explain their functional significance [87, 90, 91]. Close examination of the receptive fields of neurons in the associative cortex of the cat revealed that stimulus location is likely represented in a distributed population code and that unimodal neurons could contribute by improving the accuracy of the encoded stimulus location [87]. Furthermore, the response patterns of unimodal neurons in the cat extrastriate cortex revealed that while these neurons do not respond to their non-preferred modality, their activation is nonetheless modulated when non-preferred and preferred stimulus occur at the same time [90], further suggesting that unimodal neurons contribute to a population code. Varying the proportion of uni- and bimodal neurons might thus contribute to a continuum of multisensory responses across different cortical areas [92]. Thus, our proposed explanation for the mixture of uni- and bimodal neurons in terms of optimal decoding of input origin integrates well into the existing literature. A direct experimental test of this decoding prediction will require simultaneous recordings from V1, S1, and RL at sufficient spatial resolution. The two-photon recordings used here were acquired sequentially across areas, and the wide-field recordings, while simultaneous, do not provide cellular-resolution activity in RL. Future simultaneous multi-area recordings could therefore test whether the ability to reconstruct V1 and S1 activity from RL activity increases across development, as predicted by the model.

Together, our results suggest that structured spontaneous activity can provide a developmental scaffold for the emergence of aligned multisensory maps and efficient multimodal representations in higher-order cortex.

Methods

Experimental procedure

Animals

All experimental procedures were approved by the institutional animal care and use committee of the Royal Netherlands Academy of Sciences. C57BL/6J mice of both sexes were used. All animals were aged between postnatal days (PN) 8-16. Mice of this strain open their eyes at P14.

In utero electroporation

For wide-field imaging (Figure 2) pyramidal neurons in layer 2/3 of the visual cortex were transfected with GCaMP6s (2 mg/ml) and DsRed (2 mg/ml) at embryonic day (E) 16.5 using in utero electroporation [93]. Pregnant mice were anesthetized with isoflurane and a small incision (1.5–2 cm) was made in the abdominal wall. The uterine horns were carefully removed from the abdomen, and DNA was injected into the lateral ventricle of embryos using a sharp glass electrode. Voltage pulses (five square wave pulses, 30 V, 50-ms duration, 950-ms interval, custom-built electroporator) were delivered across the brain with tweezer electrodes covered in conductive gel. Embryos were rinsed with warm saline solution and returned to the abdomen, after which time the muscle and skin were sutured.

Surgery

Animals were anesthetized with isoflurane (3 percent in 1 l/min O2). After anesthesia had become effective, lidocaine was used for local analgesia and a head bar with an opening (Ø 4 mm) above the visual cortex (0.5–2.5 mm rostral from lambda and 1–3 mm lateral from the midline) was attached to the skull with superglue and dental cement. For calcium imaging, a small craniotomy above the visual cortex was performed. The exposed cortical surface was kept moist with cortex buffer (125 mM NaCl, 5 mM KCl, 10 mM glucose, 10 mM HEPES, 2 mM MgSO4 and 2 mM CaCl2 [pH 7.4]).

Calcium indicator application

For the imaging experiments shown in Figure 1, the calcium-sensitive dye Cal-590 (Aat Bioquest) was dissolved in 4 µl pluronic F-127, 20 percent solution in DMSO (Invitrogen) and further diluted (1:10) in dye buffer (150 mM NaCl, 2.5 mM KCl, and 10 mM HEPES) to yield a final concentration of 0.5 mM. The dye was then pressure-ejected into S1, RL and V1 at 10–12 psi for 8–13 min with a micropipette (3–5 MΩ) attached to a picospritzer (Toohey).

Image acquisition

One hour before imaging, anesthesia was stopped, and animals were recorded in the absence of anesthesia. In vivo calcium imaging with cellular resolution (Figure 1) was performed on a Nikon multi-photon microscope (A1R-MP) with a 0.8/16x water-immersion objective and a Ti:Sapphire laser (Chameleon II, Coherent) Nikon software. Pixel size was 650 nm and images of 330 by 330 µm were recorded at 5-10 Hz.

In utero electroporated pups were used for wide-field calcium imaging of visual and somatosensory cortex (Figure 2). Calcium events were recorded with a Movable Objective Microscope (MOM, Sutter Instrument). Time-lapse recordings were acquired with a 4x objective (0.8 NA, Olympus) and blue light excitation from a Xenon Arc lamp (Lambda LS, Sutter Instrument Company). A CCD camera (Evolution QEi, QImaging) was controlled by custom-made LabVIEW (National Instruments) based software and images were acquired at a frame rate of 20 Hz.

Image processing

Images were processed as described previously [47]. For two-photon image processing, drift and movement artifacts were removed from each recording using NoRMCorre [94]. Each recording was aligned to the first recording in the series to remove any movements between recording sessions. Delta F stacks were made using the mean fluorescence per pixel as baseline, F0. ROIs were hand-drawn using ImageJ (NIH). Automated transient detection and further data processing was performed using custom-made MATLAB software (MathWorks).

Epifluorescence image processing: Delta F stacks were made using the average fluorescence per pixel as baseline. V1 was identified based on activity coordinates and shape after this method of identification was confirmed through immunohistochemistry for vGluT2 [19]. Recordings were performed in an area clearly defined by lambda and the brain’s midline. Our experience shows that the field of view varies across experiments only by approximately 200 µm. To precisely overlay the field of view with the boundaries of sensory areas, we use the extent of individual network events and the organization of our topographic maps that helped identify these boundaries, in particular those of V1. We used previously published cortical area outlines as a reference map [9]). Supplementary Figure S3 shows examples of how activity patterns and topographic organization reveal the outlines of sensory cortical areas.

Statistical analysis

In two-photon recordings (Figure 1), activations within each cell were detected as in [47] as a peak with both an absolute height and relative prominence of at least 5% ΔF/F0. Participation rate was determined by % of active cells during an event out of total cells for that region. When more than 5% of recorded cells were active simultaneously, the activation was considered an event. Event duration was calculated as length of time above half max of the peak of activation. For previously published analyses of the wide-field calcium imaging data [47] (Figure 2), events were detected as calcium peaks with an amplitude above 2x the standard deviation. However, the functional correlation maps and lagged cross-correlation analyses shown in Figure 2 were computed from continuous fluorescence time courses and did not require event detection.

Functional correlation maps

Functional correlation maps were generated as in [29], from three seed areas, each 10 × 10 pixels. For the main analysis in Figure 2, seed areas were placed in RL to determine whether distinct RL locations were correlated with spatially distinct regions in V1 and S1. For the additional examples in Supplementary Figures S1 and S3, seed areas were placed in V1, S1, or RL, as indicated in each panel. For each of the three chosen seed areas, the time course of each seed area was computed by averaging pixel values within the seed region for each frame. We then computed the Pearson correlation coefficient between each seed time course and the time course of every pixel in the field of view. The three resulting correlation maps were combined into an RGB image, with each color channel corresponding to one seed location. In Figure 2E and the top rows of Supplementary Figure S1, the RGB values represent the raw correlation values for the three seed locations. In Figure 2F and the bottom rows of Supplementary Figure S1, each pixel is assigned to the color channel with the highest correlation coefficient, and pixels below the chosen correlation threshold are shown in black. Supplementary Figure S3 uses the same functional-correlation-map approach to illustrate how cortical-area outlines were assigned.

Linear mixed model

To correctly account for the nested data structure commonly encountered in neuroscience research [45], we fitted a linear mixed-effects model that treats animal and sample identity as random factors. We included interaction effects between age and region within our fixed effects. No transformations of the data were performed. Estimated effect sizes are shown in Figure 1H and Supplementary Tables S1-S3. Effect sizes on cortical region affected the intercept, but age and age-region interaction estimates affected the slope of the regression line. One-hot encoding was used because the regions are categorical variables. This means that one region, i.e. V1, was chosen as baseline and every other region was compared to this baseline. Any differences due to region were added/subtracted from the intercept for V1. Age and age-region interaction effects changed the slope and were therefore added or subtracted from the estimated slope for region V1. For Figure 1H, V1 was used as the reference region. To provide direct comparisons between all cortical areas, we also refit the same model using S1 or RL as the reference region; these results are reported in Supplementary Tables S1 and S3. Since PN8 was the earliest age used in our data, 8 was subtracted from all ages so that PN8 is the estimated intercept. Our linear model could not be valid outside of the ages used (PN8-PN16) and no extrapolation should be done without experimental confirmation. This multilevel analysis allowed us to examine the effects of age and cortical region on four response variables: amplitude, event rate, participation rate, and duration. To obtain p-values, we used Satterthwaite’s degrees of freedom method, as implemented in the lmerTest package [95] in R. The linear mixed model formula used was: (3) y=β0+β1⋅S1+β2⋅RL+(β3+β4⋅S1+β5⋅RL)(age−8)+ϵij where y is one of the response variables, i.e. amplitude, event rate, participation rate, or duration and age is postnatal day. β0 is the intercept for V1 at age PN8, and β1 and β2 are the differences in intercept for cortical region (S1 and RL are dummy variables set to 0 if not predicting that region or 1 if predicting that region). β3 is the slope for V1 over age and β4 and β5 are the differences in slope (age-region interaction effects) from V1 over age for the cortical region of interest. ϵij is the error from both random effects (mouse ID and sample ID) and residual error. When S1 or RL was used as the reference region, the same formula and age-centering were used, but the dummy variables were redefined relative to the corresponding reference area.

Rate-based network model

Extending the model introduced by [28], we studied a feedforward, rate-based network with two one-dimensional input layers, v and s, and one output layer r with N v, N s, and N r neurons each. We used periodic boundary conditions in each layer to avoid edge effects. The initial connectivity between v and r, Wv, and between s and r, Ws, was initially all-to-all with uniformly distributed weights in the range wini ∈ [a, b]. In addition, a topographic bias was introduced by modifying the initially random connectivity matrix to have the strongest connections between neurons at the matched topographic location. These biased weights decayed with a Gaussian profile with increasing distance from the matched topographic location (Figure 3), with amplitude a and spread s. Note that we used the subscripts ‘s1’ and ‘v1’ to denote which brain area the parameters referred to. In simulations in which we varied the initial S1-to-RL connectivity bias, we varied the amplitude as1 of this Gaussian bias while keeping its spread fixed. During the evolution of the weights, soft bounds were applied on the interval [0, Wmax].

The input populations v and s received spontaneous activation from a homogeneous Poisson process with an inter-event interval of 150 ms, implemented using a Binomial random variable with sufficiently small time steps of 1 ms. S1 and V1 neurons were driven by a combination of independent and shared spontaneous events. Each independent event was generated separately in either V1 or S1 and activated a randomly chosen, contiguous set of neurons. The percentage of neurons activated during an event was determined by Lrangev or LrangeS which was randomly chosen from a uniform distribution between 20% and 80% (See Table 1). The activated neurons were topographically next to each other with periodic boundary conditions. A shared event in V1 and S1 activated the identical set of contiguous neurons, topographically matched between V1 and S1. When shared events were included in the simulation, the inter-event interval in each input area was kept at 150 ms by splitting the total event rate into one shared Poisson process and two area-specific independent Poisson processes. If pshared denotes the fraction of shared events and λ = 1/150 ms−1 is the total event rate per area, shared events were generated with rate psharedλ, and independent events in each area were generated with rate (1 − pshared)λ. Thus, the total event rate experienced by each input area was psharedλ + (1 − pshared)λ = λ, while the fraction of events shared between V1 and S1 was varied. A spontaneous event increased the firing rate of a fraction of neighboring neurons to Lamp for Ldur seconds. The activity of the output layer r was given as a leaky integrator of the input populations, (4) τrr˙i=−ri+∑jN0Wijvvj+∑kN5Wikssk. We studied the weight evolution under activity-dependent learning rules. In particular, following [28] we chose a rule that combines Hebbian and heterosynaptic effects that can produce selectivity, (5) τwWikv=ri(vk−θ) and (6) τwWiks=ri(sk−θ) where θ is a heterosynaptic offset constant that defines a distance-dependent heterosynaptic depression. Heterosynaptic plasticity occurred when a nearby unstimulated synapse stabilized or depressed due to the plasticity in a stimulated one. This helped to stabilize the Hebbian plasticity [52]. Table 1 lists all parameters. For simulations with varying correlation values (Figure 3), the percentage of correlated events was chosen from a uniform distribution between 0 and 1, U[0,1]. For simulations with varying S1 bias, the Gaussian spread was fixed at 4, and the amplitude of the S1-to-RL Gaussian connectivity bias, as1, was chosen to be 0.2, 0.4 or 0.6 (Figure 4A,B), or chosen from a uniform distribution between 0.05 and 0.4 (Figure 4C-E). Additional simulations were performed to test whether S1-like activity statistics alone could account for map alignment. In these simulations, as1 was varied between 0.05 and 0.5, while S1 inter-event interval was varied between 130 and 150 ms or S1 event amplitude was varied between 0.9 and 1 (Figure S2). Simulation time of 500 s was used.

Expected weight dynamics and bimodality threshold

Since the dynamics of the output layer change much faster than the weights, τr ≪ τw, we made a steady state assumption as in our previous two-layer model [28] and computed (7) r¯i=∑jNWijvvj+∑kNWikssk. This allowed us to write the expected change in weight between (e.g.) the v and r populations as a function of the input statistics of the v and s population, (8) τw⟨W˙ikv⟩=∑jNWijv⟨vj(vk−θ)⟩+∑jNWijs⟨sj(vk−θ)⟩. Here, the expectation is taken over the input activity statistics for a fixed connectivity state. Because of the separation of timescales between activity and learning, the weights Wij are treated as fixed quantities during this calculation and can therefore be taken outside the expectation. This formulation allowed us to investigate the critical required amount of correlation between cells in v and s that permits the formation of stable bimodal cells in r. Specifically, we asked under which input correlation a depressed V1-to-RL synapse would become potentiated when the corresponding S1-to-RL synapse was already potentiated. In particular, we investigated the expected change of weight in the configuration where all connections between v and r were fully depressed and only one connection between s and r was potentiated: (9) Wiko=0 and (10) Wiks={Wmax,i=k0,otherwise. The expected weight change in this configuration was (11) τw⟨Wikv⟩=Wmax⟨si(vi−θ)⟩=Wmax(Cov(si,vi)+μvμs−μsθ) where Cov is the covariance and µv, µs are the average firing rates. The critical amount of correlation, where potentiation and depression balance each other out, is characterized by the condition (12) τw⟨W˙ikv⟩=0 which we solved for the correlation, γsv = Cov(si, vi)/(σvσs), as (13) γsv=(μsθ−μvμs)/(σvσs). Assuming that rates and variances in v and s are equal and that the activity is Poisson, µ = σ2, the condition for the critical correlation further simplifies to (14) γvs=θ−μ.

Optimal weight matrix for correlated input populations

We considered the concatenated input activity at each time point, X = [v; s], of size M × 1, where M = N v + N s input neurons. This input drives activation in the “output” population, r = ϕ(WX), where ϕ is a (possibly non-) linear transfer function, r has size N r × 1 and W has size N r × M. The decoding problem introduced in Figure 5 used a sigmoid transfer function, i.e. (15) ri=[1+exp⁡(−(ui−u¯))]−1, where ui = (WX)i and u denotes the mean preactivation across RL neurons. However, to derive the optimal weight matrix analytically, a linear transfer function was used in order to show an analytical solution for this simplified case. This decoding problem could be phrased as inferring activity of the input populations from the output population, X^=Aϕ(WX), where A is the pseudoinverse of W, A = (WT W)−1WT. We then searched for the optimal weight matrix, Wopt = [Wv, Ws] of size N r × M, where Wv represents the weights from v to r, and Ws represents the weights from s to r, that minimizes the squared error between X and X^,E=(X−X^)2 and therefore maximizes R2.

As the dimension of r is substantially smaller than the dimension of X, dim(r) ≪ dim(X), the optimal weight matrix needs to preserve the largest possible amount of variability of the input while projecting it into a lower dimensional space. Using insights from principal component analysis theory [96], we could demonstrate that the optimal weight matrix is composed of eigenvectors of the covariance matrix, ⟨XXT ⟩, with the largest eigenvalues, (16) ⟨XXT⟩Wi=λiWi, whereλ1≥λ2≥…≥λM. To better understand the structure of the eigenvectors, we could exploit the fact that in our set-up the covariance matrix, C = ⟨XXT ⟩, is circular, i.e. every row of C can be obtained by a cyclic shift of the generating vector, c = [c0, c1, …, cM−1]: (17) C=(c0c1c2⋯cM−1cM−1c0c1⋯cM−2cM−2cM−1c0⋯cM−3⋮⋮⋮⋱⋮c1c2c3⋯c0). Eigenvectors and eigenvalues of circulant matrices have well-known closed form solutions [97]. The eigenvectors are the Fourier modes: (18) vj=1M[1,ωj,ω2j,…,ω(M−1)j],forj=0,1,…,M−1. where ω = e2πi/M, a primitive M-th root of unity. The corresponding eigenvalues can be computed as (19) λj=∑k=0M−1ck⋅ωjk. Since the covariance matrix is symmetric, i.e. cj = cM−j for all j and all values are real in the covariance matrix we rewrote the eigenvalues in terms of their real parts: (20) λj=c0+2∑k=1M/2ckcos⁡(2πjkM), whenMis even. For a real-symmetric circulant matrix, the real and imaginary parts of the eigenvectors are themselves eigenvectors: (21) vj=Re(vj)=1M[cos⁡(2πj0/M),cos⁡(2πj1/M),…,cos⁡(2πj(M−1)/M)], (22) wj=Im(vj)=1M[sin⁡(2πj0/M),sin⁡(2πj1/M),…,sin⁡(2πjk(M−1)/M)]. The first M/2 eigenvectors with nonzero eigenvalues from v and w will comprise Wopt which maximizes R2. Only the first M/2 are used since the dimension of r is M/2 in our case where N v = N s and can have a maximum of M/2 eigenvectors.

Our experimental set-up with correlated input populations translates into a generating vector c that has nonzero terms around the diagonal and offset from the diagonal by M/2. If the inputs between V1 and S1 are fully correlated, then ck = 1 for k ∈ {0, …, W/2, (M − 1)/2 − W/2, …, (M − 1)/2 + W/2, (M − 1) − W/2, …, M − 1} and ck = 0 else, where W is the extent of spatial correlation of the neurons. Taking this into consideration and removing any terms that simplify to zero due to the periodicity of the cosine function gives us the eigenvalues: (23) λj={2+4∑k=1W/2cos⁡2πjkM,ifjis even,0,ifjis odd. Hence, the eigenvectors with nonzero eigenvalue of the covariance matrix are exactly the first half of the eigenvectors of the sin and cos functions with an even-numbered period. Due to this simplification of the odd j’s leading to λj = 0 and even numbered periods, this results in eigenvectors that are identical on the indices {0, …, M/2 − 1} and {M/2, …, M} which is the definition of a bimodal cell and therefore all cells must be bimodal to maximize the R2 when input correlation = 1 between V1 and S1.

Topography

Network topography was defined to measure the degree to which the initial topographical bias is preserved in the final receptive field. To calculate topography, we calculated the Pearson’s correlation coefficient, r, of the final receptive field with an ideal receptive field template (Figure 3). The ideal receptive field template that was used has weights of 1 with a Gaussian spread of 4 from center.

Alignment

Alignment was defined to measure the similarity between the final receptive fields of V1 and S1. To calculate alignment, we calculated the Pearson’s correlation coefficient, r, between the final receptive field of S1 and the final receptive field of V1.

Reconstruction

To calculate the reconstruction of V1 and S1 activity from RL in Figure 5, we fitted a linear model to predict the combined V1 and S1 activity patterns from RL activity using 1000 s of simulated activity. Duplicate RL activity patterns were removed, and 80% of the remaining data was used for fitting to avoid over-fitting. Then the coefficient of determination was calculated as (24) R2=1−SSresSStot where SSres is the sum of the residuals squared and SStot is the total sum of squares, using only one sensory area’s activity at a time, e.g. RL vs. V1 or RL vs. S1 activity compared with the linear fit of all 3 layers.